The conserved ATPase PCH-2 controls the number and distribution of crossovers by antagonizing crossover formation in C. elegans

Version 1

Abstract

Abstract

Meiotic crossover recombination is essential for both accurate chromosome segregation and the generation of new haplotypes for natural selection to act upon. While the conserved role of the ATPase, PCH-2, during meiotic prophase has been enigmatic, a universal phenotype that is observed when pch-2 or its orthologs are mutated is a change in the number and distribution of meiotic crossovers. Here, we show that PCH-2 controls the number and distribution of crossovers by antagonizing crossover formation. This antagonism produces different effects at different stages of meiotic prophase: early in meiotic prophase, PCH-2 prevents double strand breaks from becoming crossovers, limiting crossovers at sites of initial DSB formation and homolog interactions. Later in meiotic prophase, PCH-2 winnows the number of crossover-eligible intermediates, contributing to the reinforcement of crossover-eligible intermediates, designation of crossovers and ultimately, crossover assurance. We also demonstrate that PCH-2 accomplishes this regulation through the meiotic HORMAD, HIM-3. Our data strongly support a model in which PCH-2’s conserved role is to remodel meiotic HORMADs throughout meiotic prophase to destabilize crossover-eligible precursors, coordinate meiotic recombination with synapsis, and contribute to the progressive implementation of meiotic recombination, guaranteeing crossover control.

Introduction

Meiosis is a specialized type of cell division that reduces chromosome number by half, resulting in the production of four genetically diverse haploid gametes, so that fertilization during sexual reproduction restores diploidy. This process occurs in two stages: meiosis I, in which homologous chromosomes are partitioned, and meiosis II, in which sister chromatid are segregated. The regulation of meiosis is crucial for ensuring that both genetic recombination and chromosome segregation occur accurately.

During prophase I, homologous chromosomes pair and this pairing is stabilized through a process called synapsis, in which a protein structure called the synaptonemal complex (SC) holds homologs together. This close association between homologs during synapsis facilitates crossover formation, the process in which double strand breaks (DSBs) are deliberately introduced into the genome, repaired by meiosis-specific mechanisms that exchange DNA between homologous chromosomes and generate the chiasmata, or linkage, that promotes accurate meiotic chromosome segregation.

Therefore, disruptions in pairing, synapsis, or recombination prevent the formation of chiasmata and can lead to meiotic errors such as nondisjunction, resulting in aneuploid gametes. Errors in human meiosis are associated with birth defects, such as Down and Turner syndromes, infertility and miscarriages, underscoring the importance of understanding meiosis to human health (Hassold & Hunt, 2001).

In addition to the fundamental role that recombination plays in ensuring accurate chromosome segregation, crossover recombination accomplishes another important function: it generates new haplotypes for natural selection to act upon to drive evolution. Thus, to assure a random assortment of alleles on a population level, the distribution of crossovers may be as tightly regulated as their number. The significance of controlling both crossover number and distribution is clearly illustrated by the existence of mechanisms such as crossover assurance, in which every pair of homologous chromosomes gets at least one crossover; crossover homeostasis, in which the number of crossovers remains relatively invariant even if the number of recombination precursors change; and crossover interference, in which the presence of a crossover inhibits the formation of a crossover nearby (Gray & Cohen, 2016). DSBs typically vastly outnumber crossovers in most organisms and are introduced gradually throughout early prophase. Therefore, to accomplish this precise level of control, meiotic crossover recombination and the decision about which DSBs become crossover-eligible intermediates, and eventually crossovers, is implemented progressively throughout meiotic prophase (Cole et al., 2012; Joshi et al., 2015; Morgan et al., 2021; Yokoo et al., 2012).

PCH-2, also known as TRIP13 in mammals, is an evolutionarily ancient AAA-ATPase that plays a significant role in regulating meiosis across different organisms, including M. musculus (mice), S. cerevisiae (budding yeast), D. melanogaster (fruit flies), and C. elegans (worms) (Bhalla, 2023). PCH- 2 and its orthologs structurally remodel a family of proteins with HORMA domains (HORMADs) to control their function (Gu et al., 2022). HORMADs participate in a variety of signaling events and can exist in at least three structurally distinct conformations: a “closed” conformation, which they adopt when they bind a short peptide sequence in their own protein sequence or another protein (also called a closure motif) and their C-terminus wraps around this motif to stabilize the interaction; an “open” conformation when unbound, in which their C-terminus is discretely tucked against the HORMA domain; and an “extended” conformation, which is an intermediate between the two (Gu et al., 2022). PCH-2 and its orthologs convert the closed version of HORMADs to the open or extended versions, playing an important role in recycling HORMADs during signaling. During meiosis, closed versions of meiotic HORMADs assemble on chromosomes to form meiotic chromosome axes, which are essential for pairing, synapsis and recombination between homologous chromosomes (Kim et al., 2014). The remodeling of meiotic HORMADs by PCH-2 to an “open” or “extended” conformation is thought to reduce the levels of HORMADs on chromosomes (Börner et al., 2008; Cuacos et al., 2021; Lambing et al., 2015; Wojtasz et al., 2009) and/or increase their dynamic association and dissociation (Russo et al., 2023), modulating and coordinating homolog pairing, synapsis and recombination during prophase.

The conserved role of the PCH-2/HORMAD module in meiosis has been difficult to characterize, in part, we have argued, because of the evolutionary innovation that is an inherent aspect of sexual reproduction (Bhalla, 2023). Moreover, in some systems, such as budding yeast and plants (Herruzo et al., 2021; Yang et al., 2020), PCH-2 orthologs not only remodel meiotic HORMADs on meiotic chromosomes but also perform this function in the cytoplasm to make meiotic HORMADs available for their role(s) in meiotic nuclei. This dual role can complicate functional analyses, particularly in pch-2 null mutants. However, all meiotic systems exhibit defects in the number and distribution of crossovers when PCH-2 function is abrogated by mutation (Deshong et al., 2014; Joshi et al., 2009; Joyce & McKim, 2009; Lambing et al., 2015; Roig et al., 2010; Zanders & Alani, 2009). Unfortunately, the lack of a clear pattern when analyzing these defects in recombination in pch-2 mutants has contributed to an inability to develop a unified, integrated model of PCH-2 function in the field.

In C. elegans, meiotic HORMADs localize to meiotic chromosomes independently of PCH-2 (Deshong et al., 2014). Here, we exploit this system to show that PCH-2 is required to control the number and distribution of crossovers by antagonizing the formation of crossovers. This antagonism produces different consequences depending on the stage of meiotic prophase. In early meiotic prophase, PCH- 2 inhibits DSBs from becoming crossovers, ensuring that they are more widely distributed than sites of initial DSB formation and/or homolog interactions. Later in meiotic prophase, PCH-2 is responsible for winnowing the numbers of crossover-eligible intermediates on synapsed chromosomes, contributing to the reinforcement of crossover-eligible intermediates, designation of crossovers and ultimately, crossover assurance. Genetic analysis demonstrates that PCH-2’s regulation of crossover-eligible intermediates is through one of three essential meiotic HORMADs, HIM-3. Finally, we link PCH-2’s effect on early DSBs in early meiotic prophase to cell cycle stage, demonstrating that both limit early DSBs from becoming crossovers. We propose that PCH-2’s remodeling of HIM-3 on meiotic chromosomes destabilizes crossover-eligible intermediates throughout meiotic prophase, contributing to the progressive implementation of meiotic recombination to control the number and distribution of crossovers and accomplish crossover control.

Results

PCH-2 controls the number and distribution of crossovers in similar patterns on multiple chromosomes

We had previously shown that loss of PCH-2 reduced the frequency of double crossovers and genetic length of both an autosome (chromosome III) and the X chromosome, albeit not uniformly among genetic intervals (Deshong et al., 2014). To determine whether a more obvious pattern could be observed, we expanded our analysis by monitoring recombination genetically in both wildtype and pch-2 mutant animals using five single nucleotide polymorphisms (SNPs) that spanned 95% of Chromosomes, I, III, IV and the X chromosome (Figure 1). We excluded Chromosome II from our analysis because of the potential difficulty combining Hawaiian SNPs with the pch-2 mutation, which is linked to Chromosome II, and Chromosome V because of our use of the bcIs39 transgene to identify cross progeny, which may disrupt recombination on that chromosome.

PCH-2 controls the number and distribution of crossovers in similar patterns on multiple chromosomes.

Genetic analysis of meiotic recombination in wildtype and pch-2 mutants. DCO indicates double crossovers. Physical and genetic maps of Chromosome I, III, IV and the X chromosome are depicted to scale. Genetic distance is shown in centimorgans. A * indicates a p-value < 0.05, a ** indicates a p value < 0.01 and a *** indicates p-value < 0.001.

In wildtype animals, we observed multiple double crossovers, ranging from 1-13 depending on the chromosome. In a majority of these double crossovers (69%), one crossover was at the end of the chromosome where pairing and synapsis initiate, also called the Pairing Center (PC) (MacQueen et al., 2005), suggesting some relationship between where chromosomes might make initial contacts and the likelihood of double crossovers. On all chromosomes analyzed, we observed no double crossovers in pch-2 mutants and this difference was statistically significant for Chromosomes I, III and X. Moreover, there was a striking shift of crossovers to the PC end of all four chromosomes tested.

This shift in the distribution of crossovers to the PC ends of chromosomes was generally accompanied by a reduction in crossovers in the center of chromosomes. In C. elegans, the center of chromosomes are where double strand breaks are less numerous (Nadarajan et al., 2021; Yu et al., 2016) and where genes are more abundant. In the case of the X chromosome and chromosome III, recombination at the non-PC end was also reduced and appeared to more closely resemble the physical map at this end of these chromosomes. Thus, PCH-2 ensures a wider distribution of crossovers across chromosomes, away from regions that are more likely to undergo early homolog interactions (PC ends) and with more double strand breaks (both the PC and non-PC ends of chromosomes), and towards the center of chromosomes, where double strand breaks are less abundant.

PCH-2 prevents exogenous DSBs early in meiotic prophase from becoming crossovers

We previously showed that PCH-2 promotes crossover formation and crossover assurance through its regulation of the meiotic HORMAD, HIM-3, in C. elegans (Russo et al., 2023). Moreover, this promoting role seems linked to its localization to the synaptonemal complex (Patel et al., 2023).

However, the observed shift in the distribution of crossovers in pch-2 mutants suggests that PCH-2 may also play a role in inhibiting crossovers and that this role may be occurring in early meiotic prophase, when chromosomes are undergoing initial homolog interactions. Previous data from our lab had suggested this possibility, but experimental caveats limited our ability to properly interpret these experiments (Deshong et al., 2014).

To explicitly test this possibility, we took advantage of both the spatiotemporal organization of meiotic nuclei in the C. elegans germline, and the observation that nuclei travel in an assembly line process at a stereotypical pace to late pachytene (Jaramillo-Lambert et al., 2007), where we can cytologically assess crossover formation by staining for the essential crossover factor, COSA-1 (Yokoo et al., 2012). We performed irradiation experiments to introduce exogenous DSBs at different timepoints during meiotic prophase and analyzed PCH-2’s role in crossover formation (Figure 2A).

PCH-2 prevents exogenous double strand breaks from becoming crossovers in early meiotic prophase.

A. Illustration of the irradiation experiments in control and pch-2 mutants. Box indicates late pachytene, the area where GFP::COSA-1 foci are analyzed. B. Meiotic nuclei in control animals and pch-2 mutants 8 hours post irradiation stained for DAPI (magenta) and GFP::COSA-1 (green). Scale bar is 4 um. C. Fraction of meiotic nuclei with less than 6, 6, or greater than 6 GFP::COSA-1 foci in control animals(yellow) and pch-2 mutants (blue) mutants 8 hours post irradiation. D. Meiotic nuclei in control animals and pch-2 mutants 24 hours post irradiation with DAPI (magenta) and GFP::COSA-1 (green). E. Fraction of meiotic nuclei with less than 6, 6, or greater than 6 GFP::COSA-1 foci in control animals and pch-2 mutants 24 hours post irradiation. A *** indicates p- value < 0.001, and a **** indicates p-value < 0.0001.

In this experiment, we tested the effects of ionizing radiation at two distinct time points: 8 hours after irradiation, when we can monitor crossover formation in meiotic nuclei that received exogenous DSBs in mid-pachytene, and 24 hours after irradiation, when we monitor crossover formation in meiotic nuclei that received exogenous DSBs in leptotene/zygotene, also known as the transition zone in C. elegans. In order to detect crossovers, control and pch-2 mutants germlines were stained for GFP::COSA-1, which allowed us to visualize crossovers, and DAPI, which allowed us to visualize DNA. At 8 hours after irradiation, we observed a significant increase in nuclei containing less than 6 COSA-1 nuclei in pch-2 mutants (26%), when compared to control (9%) germlines (Figure 2B, 2C). We did not observe a significant number of nuclei with greater than 6 COSA-1 foci in either backgrounds. This observation is consistent with our and others’ previous findings: exogenous DSBs in mid-pachytene do not affect the number of crossovers in wildtype animals (Yokoo et al., 2012) and loss of PCH-2 leads to a decrease in crossover formation (Deshong et al., 2014).

Next, we examined crossover formation 24 hours post-irradiation (Figure 2D, 2E). For this time point, we also saw a significant increase in nuclei containing less than 6 COSA-1 foci (control: 13%, pch-2: 26%), similar to the 8 hour time point. However, in contrast to the 8 hour time point, we also observed a significant increase in nuclei with more than 6 COSA-1 foci in pch-2 mutants (wildtype: 2%, pch-2: 13%), indicating that loss of PCH-2 leads to increased crossover formation when nuclei in the transition zone get more DSBs. In other words, PCH-2 inhibits exogenous DSBs in early meiotic prophase from becoming crossovers, consistent with our analysis of recombination on multiple chromosomes.

PCH-2 prevents SPO-11-induced DSBs from becoming crossovers in early meiotic prophase

Our previous experiment indicated that PCH-2 is preventing exogenous DSBs in early meiotic prophase from becoming crossovers (Figure 2). To test whether this was also the case for programmed meiotic DSB formation, we used the auxin-inducible degradation (AID) system to remove the enzyme that is responsible for programmed meiotic DSBs, SPO-11 (spo-11::AID::3XFLAG) (Dernburg et al., 1998; Keeney et al., 1997). Since meiotic nuclei travel through the germline at about one cell per row per hour (Jaramillo-Lambert et al., 2007), we took advantage of this precise spatiotemporal resolution to test five different timepoints to determine if acute depletion of SPO-11 leads to any changes in crossover formation between control animals and pch-2 mutants (Figure 3A). We assayed crossover formation by analyzing both the number of GFP::COSA-1 foci in late pachytene and the presence of bivalents in diakinesis, which indicates the successful formation of chiasmata between homologous chromosomes. Meiotic nuclei travel from late pachytene to diakinesis in approximately 12 hours (Deshong et al., 2014).

PCH-2 prevents SPO-11-induced double strand breaks from becoming crossovers in early meiotic prophase.

A. Illustration of the SPO-11 depletion experiment to assay GFP::COSA-1 and bivalents in control animals and pch-2 mutants at different timepoints of auxin treatment. Each timepoint indicates when SPO-11 is depleted in the germline with auxin induced degradation. Boxes indicate late pachytene, the area where GFP::COSA-1 foci are analyzed, and diakinesis, where bivalents are analyzed. B. Representative images of meiotic nuclei in control animals and pch-2 mutants treated with auxin for 20 hours, stained for DAPI (magenta) and GFP::COSA-1 (green). Scale bar is 5 um. C. Number of GFP::COSA-1 foci in meiotic nuclei at different timepoints of auxin treatment in control (blue) and pch-2 mutants (yellow). Error bars represent standard error of the mean (SEM). D. Oocytes from control animals and pch-2 mutants stained for DAPI (magenta). Scale bar is 4um. E. Number of DAPI stained bodies in meiotic nuclei at different timepoints of auxin treatment in control animals and pch-2 mutants. Error bars represent SEM and a * indicates p-value < 0.05 and a *** indicates p-value < 0.001.

When we treated both control and pch-2 mutant worms for 36 hours with auxin, we observed a complete loss of GFP::COSA-1 foci (Figure 3C). When these animals were treated with auxin for 48 hours, to allow nuclei without SPO-11 an additional 12 hours to reach diakinesis, we observed 12 DAPI stained bodies, or univalents, which are the 6 homolog pairs that have not formed chiasmata (Figure 3E). Thus, we can reliably remove SPO-11 from meiotic nuclei throughout the germline and prevent crossover formation.

Having established that we can reliably remove SPO-11 with the AID system, we performed the same experiment at additional timepoints that remove SPO-11 at different stages of early meiotic prophase, monitoring crossover formation both at late pachytene and 12 hours later in diakinesis: 24 hours to monitor GFP::COSA-1 foci and 36 hours to monitor bivalent formation; 22 (GFP::COSA-1 foci) and 34 (bivalent formation) hours; 20 (GFP::COSA-1 foci) and 32 hours (bivalent formation); and 18 (GFP::COSA-1 foci) and 30 hours (bivalent formation). We also performed a control experiment, treating worms with ethanol for the 24 and 36 hour timepoints to ensure that this tagged version of SPO-11 was fully functional for crossover formation (Figure 3C, 3D).

At the 24 hour time point, when meiotic nuclei in the transition zone do not receive SPO-11-induced DSBs, we observe a range of GFP::COSA-1 foci in late pachytene in both control animals and pch-2 mutants with a majority of nuclei having 0 or 1 focus (control and pch-2 averages: 1.8 GFP::COSA-1 foci), indicating a loss of crossover formation, albeit with some heterogeneity. This was confirmed when we monitored bivalent formation at the 36 hour timepoint in both strains. This heterogeneity likely reflects that it takes 1-4 hours to significantly affect DSB formation through acute degradation of SPO-11 at this stage of meiosis in C. elegans (Hicks et al., 2022). In addition, there was no statistically significant difference in the number of GFP::COSA-1 foci between control animals and pch-2 mutants. However, we did observe a slight, statistically significant difference in the number of DAPI stained bodies between control animals (average number of DAPI stained bodies: 10.6) and pch-2 mutants (average number of DAPI stained bodies: 9.6), suggesting that even at this very early time point, we can detect a role for PCH-2 in preventing early SPO-11-induced DSBs from becoming crossovers (Figure 3D, 3E).

We continued to monitor GFP::COSA-1 and bivalent formation at successive timepoints. At the 22 hours and 34 hour timepoints, where SPO-11 depletion begins later in the transition zone, we saw an increase in the number of GFP::COSA-1 foci and a reduction in the number of DAPI stained bodies, suggesting that more DSBs are becoming crossovers. This trend continued at the 20 hour and 32 hour timepoints, eventually reaching wildtype numbers for GFP::COSA-1 and DAPI stained bodies by the 18 hour and 30 hour timepoints, respectively.

At the 22 hour and 36 hour timepoints, we did not detect any statistically significant difference in the number of GFP::COSA-1 foci or DAPI stained bodies between control animals and pch-2 mutants, even when we analyzed the number of nuclei that had greater than 6 GFP::COSA-1 foci. In contrast, while the number of GFP::COSA-1 foci was not statistically significantly different between control animals and pch-2 mutants at the 20 hour time point, the number of nuclei with greater than 6 GFP::COSA-1 foci was significantly higher in pch-2 mutants (control: 3%, pch-2: 10%, p value = 0.009, Fischer’s Exact Test), similar to our analysis of GFP::COSA-1 24 hours after irradiation. Even more strikingly, the number of DAPI stained bodies was significantly lower in pch-2 animals than control animals at the corresponding 32 hour timepoint, indicating that more homolog pairs had successfully formed chiasmata in pch-2 mutants. The inconsistency between the average number of GFP::COSA-1 foci and DAPI stained bodies at this time point can be explained by our ability to more reliably detect pch-2’s loss of crossover assurance when quantifying GFP::COSA-1 foci than when analyzing bivalent formation (Deshong et al., 2014; Russo et al., 2023). No difference in GFP::COSA- 1 foci or bivalent formation was observed at the 18 hour and 30 hours timepoints, respectively. These data indicate that during a narrow time frame of early meiotic prophase, likely during leptotene/zygotene given the delay in limiting DSB formation with auxin-induced degradation of SPO- 11, PCH-2 prevents SPO-11-induced DSBs from becoming crossovers and chiasmata.

PCH-2 is required for timely loading and removal of MSH-5 on meiotic chromosomes through its regulation of HIM-3

After DSB formation, a subset of DSBs are repaired through a pro-crossover pathway and eventually winnowed to a stereotypical number of crossovers. In C. elegans, this can be visualized by the loading of the pro-crossover factor MSH-5, a component of the meiosis-specific MutSγ complex that stabilizes crossover-specific DNA repair intermediates called joint molecules (Janisiw et al., 2018; Kelly et al., 2000; Snowden et al., 2004). GFP::MSH-5 begins to form a few foci in the transition zone, becoming more numerous in early pachytene before decreasing in number in mid pachytene to ultimately colocalize with COSA-1 marked sites in late pachytene (Janisiw et al., 2018; Woglar & Villeneuve, 2018; see control in Figure 4A, 4B). The mechanism through which MSH-5 functions with the SC and other meiosis factors to ultimately form crossovers has not been established but may involve phosphorylation by cyclin-dependent kinases (Haversat et al., 2022; Zhang et al., 2021).

PCH-2 is required for timely loading and removal of MSH-5 on meiotic chromosomes through its regulation of HIM-3.

A. Representative images of nuclei in different stages of meiotic prophase in control animals and pch-2 mutants stained for DAPI (magenta) and GFP::MSH-5 (green) Scale bar is 5 um. B. Scatter plot showing average GFP::MSH-5 foci per row of germline nuclei in control animals (yellow) and pch-2 mutants (blue) from the transition zone to late pachytene, normalized to 100. The line represents a rolling average of four rows. C. Representative images of nuclei in different stages of meiotic prophase in him-3R93Y mutants (left) and pch-2;him-3R93Ydouble mutants (right), stained for DAPI (magenta) and GFP::MSH-5 (green). D. Scatter plot showing average GFP::MSH-5 foci per row in him-3R93Y(brown) and pch-2;him-3R93Y mutants (pink) from the transition zone to late pachytene, normalized to 100. The line represents a rolling average of four rows. Similar data is provided for a control germline (opaque yellow) for comparison. E. Representative images of meiotic nuclei in control animals and pch-2 mutants stained for DAPI (blue), GFP::MSH-5 (green), and OLLAS::COSA-1 (red). Yellow circles indicate GFP::MSH-5 without OLLAS::COSA-1. Scale bar is 4um. F. Scatter plot showing average GFP::MSH-5 (green) and OLLAS::COSA-1 (red) foci per row in the last five rows of the germline in control animals and pch-2 mutants. The line represents a rolling average of 2 rows. G. Swarm plot showing number of GFP::MSH-5 foci in control (yellow) and pch-2 mutant (blue) nuclei with less than 6 OLLAS::COSA-1 foci (left) and 6 OLLAS::COSA-1 foci (right). Error bars represent SEM. A * indicates a p-value less than 0.05 and a ** indicates a p-value < 0.01.

Given that we have shown that PCH-2 prevents early DSBs from becoming crossovers (Figures 2 and 3), we tested whether we could cytologically detect this inhibition at the level of MSH-5 behavior in control animals and pch-2 mutants. We generated wildtype and pch-2 mutants with GFP::MSH-5 and quantified the average number of GFP::MSH-5 foci per row of nuclei from the transition zone to late pachytene (Figure 4A, 4B). We observed a statistically significant increase in the average number of GFP::MSH-5 foci per row in the transition zone in pch-2 mutants (Figure 4A, 4B, p value < 0.0001, Student’s t-test), indicating that PCH-2 typically limits MSH-5 loading at this early stage of meiotic prophase. The average number of GFP::MSH-5 foci per row were similar in number in early pachytene in both backgrounds. However, we also observed a substantial increase in the average number of GFP::MSH-5 foci per row in pch-2 mutants in the mid and late pachytene regions (Figure 4A, 4B), indicating that PCH-2 also promotes the turnover or maturation of GFP::MSH-5 foci at these later stages of meiotic prophase. Thus, PCH-2 prevents the loading of the crossover-promoting factor, MSH-5, during early meiotic prophase, consistent with both our genetic analysis of recombination (Figure 1) and its role in preventing early DSBs from becoming crossovers (Figures 2 and 3).

However, PCH-2 also appears to limit the number of crossover-eligible intermediates in mid and late pachytene, an unexpected observation given that the most severe recombination defect in pch-2 mutants is the loss of crossover assurance (Deshong et al., 2014).

We previously showed that PCH-2 genetically interacts with meiotic HORMADs to effectively control different aspects of meiosis: PCH-2 regulates pairing and synapsis through its regulation of HTP-3 and crossover recombination through its regulation of HIM-3 (Russo et al., 2023). To determine which meiotic HORMAD PCH-2 might be regulating to affect GFP::MSH-5’s loading and removal on meiotic chromosomes, we constructed gfp::msh-5;htp-3H96Y, gfp::msh-5;pch-2;htp-3H96Y, gfp::msh-5;him-3R93Y and gfp::msh-5;pch-2;him-3R93Y mutants and quantified GFP::MSH-5 foci throughout the germline.

Both htp-3H96Y and him-3R93Y mutants showed a drastic increase in the average number of GFP::MSH- 5 per row of nuclei throughout the germline (Figures 4C and 4D, Supplemental Figure 1). For example, compared to control germlines that peaked at approximately an average of 20 MSH-5 foci per row in early pachytene and then decreased in mid-pachytene, him-3R93Y germlines had closer to an average of 30 GFP::MSH-5 foci per row in early pachytene. Moreover, the average number of GFP::MSH-5 foci per row increased further in mid and late pachytene, achieving peaks closer to an average of 35 foci (Figure 4D). These very high numbers of GFP::MSH-5 were striking but do not produce an increased number of crossovers in htp-3H96Y or him-3R93Ysingle mutants (Russo et al., 2023), raising the possibility that not all of them reflect functional, crossover-specific intermediates, similar to what has been observed in the absence of synapsis (Woglar & Villeneuve, 2018). These data also suggest that these mutations, which disrupt the ability of meiotic HORMADs to adopt the “closed” conformation (Russo et al., 2023), affect the behavior of GFP::MSH-5 on meiotic chromosomes.

Quantification of GFP::MSH-5 foci in pch-2;htp-3H93Ydouble mutants was similar to htp-3H93Y single mutants, suggesting that PCH-2’s effect on the behavior of GFP::MSH-5 foci was not through its regulation of HTP-3 (Supplemental Figure 1). In stark contrast to pch-2;htp-3H93Ydouble mutants, the pch-2;him-3R93Y double mutant showed a dramatic decrease in the overall average of MSH-5 foci per row throughout the germline (Figure 4C, 4D). These averages fell between the averages in control and pch-2 mutant animals, particularly in the transition zone and late pachytene (Figure 4D), consistent with our previous report that this double mutant has fewer defects in crossover formation than either single mutant (Russo et al., 2023). These data indicate that PCH-2’s effect on promoting the removal of GFP::MSH-5 in both the transition zone and mid to late pachytene is through its regulation of HIM- 3.

pch-2 single mutants exhibit a loss of crossover assurance (Deshong et al., 2014). However, our quantification of GFP::MSH-5 in pch-2 single mutants in mid to late pachytene showed an increased number of crossover-eligible intermediates (Figure 4A, 4B), suggesting a complex relationship between having too many crossover-eligible intermediates and crossover assurance in C. elegans. To address this inconsistency, we generated strains with both OLLAS::COSA-1 (Janisiw et al., 2018) and GFP::MSH-5 in wildtype animals and pch-2 mutants. Consistent with our and others’ previous findings, we observe a gradual reduction in the average number of GFP::MSH-5 foci per row of nuclei in control animals as these nuclei approach the end of pachytene, ultimately converging with the average number of OLLAS::COSA-1 foci per row (Figure 4E, 4F) (Janisiw et al., 2018; Woglar & Villeneuve, 2018). In addition, we consistently observed co-localization of GFP::MSH-5 and OLLAS::COSA-1 foci in control animals, with a few MSH-5 foci persisting without COSA-1 (Figure 4E, 4F).

By contrast, we detected higher average numbers of GFP::MSH-5 foci per row across all late pachytene nuclei in pch-2 mutants and this average never converged upon the average number of OLLAS-1::COSA-1 foci per row (Figure 4E, 4F). To test whether there was a correlation between the number of GFP::MSH-5 foci and the loss of crossover assurance in pch-2 mutants, we determined the average number of GFP::MSH-5 foci in meiotic nuclei with 6 OLLAS::COSA-1 foci and those with less than 6 OLLAS::COSA-1 foci in both control and pch-2 mutant animals (Figure 4E, 4G, Supplemental Figure 2). Indeed, we observed that pch-2 mutant nuclei with less than 6 OLLAS::COSA-1 foci had significantly higher numbers of GFP::MSH-5 foci, compared to both control and pch-2 mutant nuclei with 6 OLLAS::COSA-1 foci (Figure 4E, 4G), suggesting that the inability to reduce the number of crossover-eligible intermediates in pch-2 mutants, counterintuitively, contributes to the observed loss of crossover assurance.

PCH-2 is removed from the synaptonemal complex when crossovers are designated

We have previously shown that PCH-2 localization to the SC is extended when there are partial defects in synapsis or changes in karyotype, producing an increase in crossovers and a loss of crossover interference (Deshong et al. 2014; Patel et al., 2023). These data led us to hypothesize that PCH-2’s presence on chromosomes promotes crossover formation. However, our quantification of GFP::MSH-5 indicates that PCH-2 is required to limit the number of crossover-eligible intermediates, in direct contrast to our proposed hypothesis (Figure 4). Therefore, we decided to revisit what role PCH-2 localization to the SC might play in regulating crossover formation. To this end, we localized PCH-2 in dsb-2 mutants, in which DSB formation is substantially reduced and fewer crossovers are formed in C. elegans (Rosu et al., 2013). We performed this experiment because, in budding yeast, similar mutants that reduce DSB formation rely on Pch2 to successfully complete meiosis (Joshi et al., 2009; Zanders & Alani, 2009).

Because the recombination defect in dsb-2 mutants worsens with age (Rosu et al., 2013), we looked at PCH-2 localization in both young (24 hours post L4 larval stage) and older animals (48 hours post L4). We observed a unique and striking localization pattern in dsb-2 mutants that we had not observed before. When stained for PCH-2 and GFP::COSA-1, nuclei retain PCH-2 onto chromosomes far into late pachytene, past the normal region when PCH-2 typically is removed from chromosomes (Figure 5A, 5B). However, unlike what we have observed in other mutants that have defects in synapsis or changes in karyotype, this retention of PCH-2 is not uniform among all late pachytene nuclei in dsb-2 mutants. Instead, most nuclei lose PCH-2 localization in late pachytene while some retain it. We tested whether there was a relationship between the retention of PCH-2 and the number of GFP::COSA-1 foci and found that 96% of nuclei that lose PCH-2 have at least 1 GFP::COSA-1 focus (Figure 5A, 5B). In older animals (see 48 hours post-L4), this pattern was even more clear: 77% of nuclei without PCH-2 had 1 or more GFP::COSA-1 focus and 93% of meiotic nuclei without a GFP::COSA-1 focus retained PCH-2.

PCH-2 is removed when crossovers are designated.

A. Representative images of meiotic nuclei in dsb-2 animals 24 hours post L4 and 48 hours post L4 stained for DAPI (magenta), PCH-2 (red), and GFP::COSA-1 (green). Scale bar is 4 um. B. Stacked histograms showing percentage of PCH-2 positive and negative nuclei with (lime) and without (dark green) GFP::COSA-1 foci in dsb-2 mutants at 24 hours post L4 and 48 hours post L4. C. Representative images of meiotic nuclei in dsb-2::AID and dsb-2::AID;pch-2 mutants treated with auxin and stained for DAPI (magenta) and GFP::COSA-1 (green). Scale bar is 5 um. D. Swarm plot showing the number of GFP::COSA-1 foci in dsb-2::AID (gray) and dsb-2::AID;pch-2 (lemon) mutants when treated with ethanol or auxin. Error bars represent the SEM. A ** indicates a p-value less than 0.01 and a **** indicates p-value < 0.0001.

These data indicate that PCH-2’s removal from the SC is in response to, coincident with or facilitates crossover designation. To distinguish between these possibilities, we analyzed GFP::COSA-1 foci in dsb-2::AID (Zhang et al., 2018) and dsb-2::AID;pch-2 worms, reasoning that if PCH-2’s removal is in response to or coincident with designation, we should not detect any differences in the number of GFP::COSA-1 foci in dsb-2::AID and dsb-2::AID;pch-2 worms. We performed these experiments with the auxin-inducible degradation system because of variability in the dsb-2 mutant background that affected reproducibility of our experiments.

Upon auxin treatment of dsb-2::AID worms, we observed a statistically significant (p value < 0.0001, Mann-Whitney U test) decrease in the number of GFP::COSA-1 foci, in comparison with ethanol- treated worms, verifying that we can reliably knock down DSB-2 with the AID system (Figure 5D).

When we performed the same experiment in dsb-2::AID;pch-2 worms, the average number of GFP::COSA-1 foci further decreased, indicating that fewer crossovers are designated when DSBs are reduced and PCH-2 is absent (Figure 5C, 5D). These data suggest that PCH-2’s presence on the SC prevents crossover designation. Given the correlation between elevated GFP::MSH-5 foci and the loss of crossover assurance we observe in pch-2 mutants, we propose that PCH-2 is retained on meiotic chromosomes to ensure that extra crossover-eligible intermediates are removed and crossover designation is delayed until crossover assurance can be guaranteed in C. elegans.

PCH-2 and high CHK-2 activity control the fate of early double strand breaks

In C. elegans, meiotic cell cycle entry and progression depends on the activity of CHK-2, the meiosis- specific ortholog of the DNA-damage kinase Chk2/CHEK2 (Baudrimont et al., 2022; Castellano-Pozo et al., 2020; Kim et al., 2015; MacQueen & Villeneuve, 2001; Zhang et al., 2023). Meiotic nuclei in leptotene/zygotene are characterized by high CHK-2 activity, which drops to intermediate activity in mid-pachytene (Kim et al., 2015; Zhang et al., 2023). In late pachytene, CHK-2 activity is inactivated by the recruitment of polo-like kinase, PLK-2, to the SC, which enables crossover designation (Zhang et al., 2023). Thus, CHK-2 activity also has consequences for the progression of meiotic recombination and DNA repair.

Given that early DSBs are prevented from becoming crossovers by PCH-2 in early meiotic prophase, we wanted to test if high CHK-2 activity in leptotene/zygotene also contributed to the fate of these early DSBs. To evaluate this possibility, we used syp-1T452A mutants (Sato-Carlton et al., 2018). The SC component, SYP-1, is phosphorylated by CDK-1 on T452, producing a Polo box binding motif that recruits PLK-2 to the SC, contributing to the inactivation of CHK-2 when chromosomes are synapsed (Brandt et al., 2020; Zhang et al., 2023). Therefore, in syp-1T452A mutants, CHK-2 activity remains high throughout most of meiotic prophase in the C. elegans germline (Figure 6A).

PCH-2 and high CHK-2 activity control the fate of early double strand breaks.

A. Illustration of CHK-2 activity in wildtype and syp-1T452A germlines. B. Representative images of meiotic nuclei late pachytene in syp-1T452A and pch-2;syp-1T452A mutants stained for DAPI (magenta) and GFP::COSA-1 (green). Scale bar is 5 um. C. Swarm plot showing number of GFP::COSA-1 foci in control animals (blue), pch-2 (yellow), syp-1T452A (maroon), and pch-2;syp-1T452A (light blue) mutants. Error bars represent SEM. D. Oocytes from syp-1T452A and pch-2;syp-1T452Amutant worms stained for DAPI (magenta). Scale bar is 4um. D. Swarm plot showing number of DAPI stained bodies in control animals (blue), pch-2 (yellow), syp-1T452A (maroon), and pch-2;syp-1T452A (light blue) mutants. Error bars represent SEM. A **** indicates p-value < 0.0001.

We quantified the total number of GFP::COSA-1 foci in late pachytene nuclei in syp-1T452A mutants and observed a significant decrease in the average number of GFP::COSA-1 foci when compared to control animals (Figures 6B and 6C), consistent with previous findings that syp-1T452A mutants delay designation and have fewer crossovers (Zhang et al., 2023). These data are also consistent with the possibility that some DSBs fail to become crossovers when CHK-2 activity remains high. If this hypothesis is correct, we predict that the combination of high CHK-2 activity and loss of pch-2 should produce more crossovers. We generated pch-2;syp-1T452A double mutants to test this hypothesis.

When we quantified GFP::COSA-1 in pch-2;syp-1T452A double mutants, we detected a significant increase in the average number of GFP::COSA-1 foci compared to syp-1T452Asingle mutants (Figure 6C), in strong support of our hypothesis.

To further verify that the pch-2 mutation suppresses the crossover defect in syp-1T452Amutants, we also monitored bivalent formation in diakinesis nuclei. syp-1T452A single mutants exhibit an average of 7.12 DAPI staining bodies in diakinesis, higher than both control animals and pch-2 single mutants (Figure 6D). By contrast, pch-2;syp-1T452A mutants had an average of 6.05 DAPI staining bodies (Figure 6D, 6E), directly supporting our COSA-1 analysis and indicating that both PCH-2 function and high CHK-2 activity collaborate to control the fate of DSBs and prevent some of them from becoming crossovers in early meiotic prophase.

Discussion

We have shown that PCH-2 antagonizes crossover formation throughout meiotic prophase and that this regulation occurs through one of the three essential meiotic HORMADs in C. elegans, HIM-3 (Figure 4). We propose that PCH-2 remodels HIM-3 on meiotic chromosomes to destabilize crossover-eligible intermediates, thus limiting which DSBs will become crossovers (Figure 7). However, this antagonism has different consequences depending on when during meiotic prophase it occurs, underscoring the importance of temporal regulation of these events. During leptotene/zygotene, when CHK-2 activity is high (Kim et al., 2015; Zhang et al., 2023) and PCH-2 is present as foci on chromosomes (Deshong et al., 2014), PCH-2 prevents crossover formation at some sites of initial DSB formation and early homolog interactions (Figure 7B). In this way, PCH-2 promotes a wider distribution of crossovers across the genome (Figure 1). In pachytene, when CHK-2 activity has decreased (Kim et al., 2015; Zhang et al., 2023) and PCH-2 is localized to the SC (Deshong et al., 2014), PCH-2 winnows crossover-eligible intermediates, ensuring their reinforcement, designation and crossover assurance (Figure 7B). When there are defects in recombination, such as partial synapsis (Deshong et al., 2014), changes in karyotype (Patel et al., 2023) or too few DSBs (Figure 5), PCH-2 persists on the SC to prevent designation, guarantee crossover assurance and some degree of homeostasis, independent of an additional feedback mechanism that increases DSB formation (Patel et al., 2023). That persistence of PCH-2 also disrupts crossover interference in two of these situations (Deshong et al., 2014; Patel et al., 2023) strongly suggests that crossover interference is mechanistically linked to assurance and homeostasis.

PCH-2 remodels HIM-3 to disassemble crossover-eligible intermediates, controlling crossover distribution and number.

A. Model for how pch-2 and him-3R93Ymutations genetically interact to affect the progression of meiotic recombination. HIM-3 adopts the closed conformation upon binding an interacting protein with a closure motif and its conversion to the extended conformation is facilitated by PCH-2’s remodeling of its HORMA domain. B. Model for how PCH-2 and HIM-3 progressively implement meiotic recombination during different stages of meiotic prophase.

him-3R93Y;pch-2 double mutants have stronger crossover assurance than either single mutant but less than wildtype animals (Russo et al., 2023), a phenotype which can now be explained by the behavior of GFP::MSH-5 foci in these double mutants (Figure 4C, 4D). We have shown that HIM-3R93Y mutant protein can adopt the closed conformation and loads on meiotic chromosomes similar to wildtype HIM- 3 (Russo et al. 2023). In vitro analysis shows that HIM-3R93Y binds its closure motif with reduced affinity, likely affecting its ability to adopt the closed conformation in vivo (Figure 7A and Russo et al. 2023). The proposed role of PCH-2 in meiosis is to remodel meiotic HORMADs from the closed conformation to the extended one (Figure 7A), regulating their association with chromosomes (Bhalla, 2023). However, since meiotic HORMADs are not visibly depleted from chromosomes during meiotic progression in C. elegans (Couteau et al., 2004; Couteau & Zetka, 2005; Goodyer et al., 2008; Martinez-Perez & Villeneuve, 2005), this genetic interaction supports a role for PCH-2 in temporarily reducing the occupancy of meiotic HORMADs on meiotic chromosomes, destabilizing interactions with partner proteins that modulate the progression and fidelity of meiotic recombination (Russo et al., 2023). Thus, PCH-2’s remodeling of HIM-3 would disrupt protein-protein interactions that underlie crossover-eligible intermediates, destabilizing them and reducing their number on chromosomes, contributing to crossover control (Figure 7B). Since meiotic HORMADs are essential for meiotic chromosome axis structure and function, our data therefore support a role for the meiotic axis in crossover control, as suggested by previous reports (Chu et al., 2024; Girard et al., 2023; Lambing et al., 2020; Nabeshima et al., 2004). However, the disassembly of crossover-eligible intermediates would also release pro-crossover factors present at these sites to concentrate at other, more stable sites on the SC, contributing to reinforcement and designation during synapsis (Girard et al., 2023; Yokoo et al., 2012). Moreover, these results raise the possibility that this behavior of PCH-2 and/or meiotic HORMADs might be regulated by post-translational modifications associated with crossover control, such as ubiquitination, SUMOylation and phosphorylation (Gray & Cohen, 2016).

Given that mutation of PCH-2 produces changes in the number and distribution of crossovers across multiple model systems, we argue that controlling crossover distribution and number is the conserved role of PCH-2 in meiosis (Bhalla, 2023). We have previously proposed that the conserved role of PCH- 2 is to coordinate meiotic recombination with synapsis (Bhalla, 2023). With these results, we further refine this model. Early in meiotic prophase, PCH-2 remodels meiotic HORMADs to prevent some DSBs from becoming crossover-eligible intermediates, widening the recombination landscape beyond early homolog interactions and/or sites that tend to be more favorable for DSB formation, also known as ‘hot spots.” For example, this may explain the localization of TRIP13, the PCH-2 ortholog in mice, to telomeres (Chotiner et al., 2024), sites that experience early homolog interactions due to the organization of meiotic chromosomes in the bouquet formation (Scherthan et al., 1996). This antagonism may also expand the regions of the genome that initiate synapsis in organisms that use DSBs to accomplish this event. This possibility is supported by the observation that loss of TRIP13 in mammals produce meiotic chromosomes that exhibit partial asynapsis (Roig et al., 2010), particularly near regions that may act as barriers to SC polymerization (Brown et al., 2005; Roig et al., 2010).

Limiting which DSBs becomes crossover-eligible intermediates in early meiotic prophase also ensures that meiotic recombination overlaps with synapsis, either completely, as in C. elegans (Yokoo et al., 2012) or partially, as in budding yeast, plants and mice (Capilla-Perez et al., 2021; Cole et al., 2012; Joshi et al., 2015; Morgan et al., 2021). Once synapsis is complete, PCH-2 continues to remodel meiotic HORMADs on chromosomes to control the gradual implementation of crossover number and distribution, reinforcing the important role that synapsis plays in mediating crossover control (Durand et al., 2022; Libuda et al., 2013).

Unexpectedly, the inability to reduce the number of crossover-eligible intermediates in pch-2 mutants, as visualized by GFP::MSH-5 foci, does not produce extra crossovers but a loss of crossover- assurance in C. elegans (Figure 4), a somewhat counterintuitive result. One interpretation of these data is that crossover-eligible intermediates may be more numerous but absent from some chromosomes in pch-2 mutants, explaining the loss of crossover assurance. Since the absence of crossover intermediates in C. elegans is accompanied by premature desynapsis of individual chromosomes (Machovina et al., 2016; Pattabiraman et al., 2017) and chromosomes in pch-2 mutants delay desynapsis (Deshong et al., 2014), we do not favor this interpretation. Instead, we propose that having too many crossover-eligible intermediates can be as deleterious to crossover assurance as having too few (Figure 7B). This possibility is further supported by the loss of crossover assurance we detect in irradiated wildtype worms, which is exacerbated in pch-2 mutants (Figure 2).

This phenomenon, where crossover-eligible intermediates need to be winnowed to some threshold number to ensure crossover assurance, may also explain the loss of crossover assurance also observed in Trip13 deficient mice (Roig et al., 2010) and on small chromosomes in budding yeast (Chakraborty et al., 2017). Alternatively, the counterintuitive relationship between the number of crossover-eligible precursors and crossover assurance in pch-2 mutants we observe might reflect an additional layer of regulation during crossover formation specific to C. elegans. Since C. elegans chromosomes are holocentric, crossovers play an additional role organizing chromosomes for the ordered release of sister chromatid cohesion during meiosis I (Martinez-Perez et al., 2008; Nabeshima et al., 2005) and extra crossovers can be deleterious to accurate chromosome segregation (Hollis et al., 2020). By contrast, in Arabidopsis, a system that appears to be able to tolerate an extraordinarily high number of crossovers with little to no effect on chromosome segregation (Durand et al., 2022), PCH2’s inability to localize to the SC produces an increase in crossover formation (Figure 7), as visualized by both MLH1 foci and the formation of chiasmata (Yang et al., 2022). Once again, an overarching theme that becomes apparent in our model is that PCH-2 may play a common role in different systems, with dramatic variations in phenotypic consequences given species-specific requirements and constraints.

We were not surprised to see high numbers of double crossovers on almost every chromosome in our genetic analysis of recombination in wildtype worms, given our previous analysis (Deshong et al., 2014). However, we were surprised to see that the majority of them were found near Pairing Centers and sites of synapsis initiation, suggesting a relationship between early homolog interactions and the formation of double crossovers. When we revisited our previous data, we observed similar patterns. In addition, we do not detect these double crossovers cytologically in C. elegans, even when using the OLLAS::COSA-1 reporter, which has been reported to identify double crossovers in spermatogenesis not visualized by GFP::COSA-1 (Cahoon et al., 2023). Crossovers that are cytologically marked by COSA-1 are known as Class I crossovers, which depend on pro-crossover factors such as MSH-5 and ZHP-3, and exhibit crossover control (Gray & Cohen, 2016). These data raise the intriguing possibility that these double crossovers are the product of the alternate, Class II, pathway of crossover formation, which relies on a different suite of proteins, does not respond to crossover control and contributes to varying degrees in different model systems (Gray & Cohen, 2016; Youds et al., 2010). Thus, important corollaries of our model may be that PCH-2 specifically coordinates recombination with synapsis to promote Class I crossovers, limit Class II crossovers and that Class II crossovers are more likely to form early in meiosis, prior to synapsis. Therefore, variations in the contribution of the Class II crossover pathways to crossover recombination and the degree of cross-talk between Class I and Class II pathways among model systems might reflect the degree to which crossover formation overlaps with synapsis (Bhalla, 2023; Gray & Cohen, 2016; Yokoo et al., 2012). Furthermore, these corollaries are entirely consistent with PCH-2’s absence from the genome of fission yeast (Wu & Burgess, 2007), a system in which chromosomes do not undergo synapsis, crossovers do not exhibit interference and all crossovers are dependent on the Class II pathway (Hollingsworth & Brill, 2004).

Our results also have some additional important implications about the regulation of DSB formation across the genome in C. elegans. We explain the shift in the recombination landscape away from the central regions of chromosomes and toward PC ends in pch-2 mutants (Figure 1) as a result of early DSBs becoming crossovers at the expense of later DSBs, suggesting that when DSBs happen in meiotic prophase affects where they happen in the genome. Specifically, we propose that chromosome arms, which are gene poor, receive DSBs early during (or even throughout) DSB formation and the center of chromosomes, which are gene rich, receive DSBs later. A similar regulation of the timing of DSB formation has been demonstrated in budding yeast, where the DSB landscape across the whole genome expands when time in prophase is increased (López Ruiz et al., 2024) and small, highly recombinogenic, chromosomes, get more DSBs later in meiotic prophase (Murakami et al., 2020; Subramanian et al., 2019). However, this temporal regulation has not been reported previously in C. elegans and suggests that this phenomenon is more widely conserved. This expansion of the DSB landscape in C. elegans to include the center regions of chromosomes later in meiosis may be a deliberate attempt for recombination to create new haplotypes for evolution to act on, despite the relative paucity of DSBs and the observation that they can result in chromosome missegregation (Altendorfer et al., 2020).

Finally, our work raises some important questions about the functional role(s) of DSBs in meiosis, aside from their contributions to crossover formation. Hicks and colleagues were the first to report that early DSBs do not contribute to crossover formation in C. elegans (Hicks et al., 2022). Here we show that these early DSBs are prevented from becoming crossovers by both PCH-2 activity and cell cycle stage, specifically in leptotene/zygotene, when homologs are initiating pairing and synapsis. In budding yeast, similar, early-occurring DSBs have been characterized as “scout DSBs” because of their preference for repair from sister chromatids, versus homologous chromosomes, and their proposed role in contributing to homolog pairing (Borde & de Massy, 2015; Joshi et al., 2015). The homolog bias that these “scout DSBs” do display seems dependent on budding yeast PCH2 (Joshi et al., 2015) but interpreting this experiment is complicated by the fact that Pch2 is also required to make the budding yeast meiotic HORMAD, Hop1, available for its loading onto meiotic chromosomes (Herruzo et al., 2021).

In contrast to budding yeast, C. elegans does not rely on DSBs to promote homolog pairing and initiate synapsis (Dernburg et al., 1998); in worms, cis-acting sites called Pairing Centers are essential for homolog pairing and synapsis (MacQueen et al., 2005). There is certainly some support in the literature for DSBs playing a role in supporting pairing and synapsis in C. elegans (Guo et al., 2022; Mlynarczyk-Evans et al., 2013; Roelens et al., 2015). However, if these early DSBs were contributing to pairing and synapsis, we would expect to see a genetic interaction between htp-3H96Y and pch-2 mutations in the installation of GFP::MSH-5 in the transition zone; we previously reported that htp- 3H96Y suppresses the acceleration of pairing and synapsis of pch-2 mutants, particularly when Pairing Center function is compromised (Russo et al., 2023). Instead, we favor the possibility that these early DSBs are generated to amplify the signaling of DNA damage kinases to prime them for their role in recombination, a role that has also been proposed for “scout” DSBs (Joshi et al., 2015). That ATM-1, a conserved DNA damage kinase that is downstream of CHK-2 in C. elegans, relies on DSBs for full activity, supports this proposed, conserved role (Yu et al., 2023).

Our work provides an important framework to finally understand the role of PCH-2 in controlling the number and distribution of crossovers, a role that we argue is its conserved role. While specific details may vary across systems, we propose that PCH-2 remodels meiotic HORMADs throughout meiotic prophase to destabilize crossover-eligible precursors, coordinating meiotic recombination with synapsis, contributing to the progressive implementation of meiotic recombination and guaranteeing crossover assurance, interference and homeostasis.

Methods and Materials

C.elegans Genetics and Genome Engineering

The C. elegans Bristol N2 was used as the wild-type strain. All strains were maintained at 20°C under standard conditions unless stated. Mutant combinations were generated by crossing. The following mutants and rearrangements were used:

LGII: pch-2(tm1458), meIs8 ([pie-1p::GFP::cosa-1 + unc-119(+)], dsb-2(me96), dsb-2(ie58[dsb- 2::AID::3xFLAG])

LGIII: cosa-1(ddr12[OLLAS::cosa-1]); htp-3(vc75)

LGIV: him-3(blt9), spo-11(ie59[spo-11::AID::3xFLAG]), msh-5[ddr22(GFP::msh-5)], ieSi38 [sun- 1p::TIR1::mRuby::sun-1 3’UTR + Cbr-unc-119(+)]

LGV: syp-1(icm85[T452A]), nT1[qIs51], bcIs39 (Plin-15::ced-1::GFP)

Genetic analysis of Recombination

The wildtype Hawaiian CB4856 strain (HI) and the Bristol N2 strain were used to assay recombination between single nucleotide polymorphisms (SNPs) on Chromosomes I, III, IV and X (Wicks et al., 2001; Bazan et al., 2011). The SNPs, primers, enzymes used for restriction digests and expected fragment sizes are included in Supplemental Table 1. To measure wild-type recombination, N2 males containing bcIs39 were crossed to Hawaiian CB4856 worms. Cross-progeny hermaphrodites were identified by the presence of bcIs39 and contained one N2 and one CB4856 chromosome.

These were assayed for recombination by crossing with CB4856 males containing myo-2::mCherry. Cross-progeny hermaphrodites from the resulting cross were isolated as L4s, and then cultured individually in 96- well plates in liquid S-media complete supplemented with HB101. Four days after initial culturing, starved populations were lysed and used for PCR and restriction digest to detect N2 and CB4856 SNP alleles. For recombination in pch-2 mutants, strains homozygous for the CB4856 background of the relevant SNPs were created, then mated with pch-2; bcIs39. Subsequent steps were performed as in the wild-type worms.

Immunostaining

DAPI staining and immunostaining was performed as in (Russo et al. 2023), 20 to 24 hours post L4 unless otherwise noted. For analyzing bivalents, the same protocol was implemented with the exception that hermaphrodites were dissected and DAPI stained 48 hours post late L4 stage, unless otherwise noted.

The following primary antibodies were used at the indicated dilutions: alpaca anti-GFP Booster (ChromoTek, gb2AF488) was used at 1:1000; rat anti-OLLAS (Invitrogen, PIMA516125) was used at 1:1000; and rabbit anti PCH-2 (Bhalla 2014) was used at 1:500. The following secondary antibodies were used at the indicated dilutions: anti-rabbit Cy3 (Jackson Labs) was used at 1:500 and anti-rat Cy5 (Jackson Labs) was used at 1:500

Irradiation Experiments

Control and pch-2 mutant L4’s were aged 12-14 hours before being exposed to 1,000 rad (10 Gy) of X-ray radiation using a Precision MultiRad 160 X-irradiator (Precision X-Ray Inc.). Germlines were then fixed and stained, 8 hours and 24 hours post irradiation.

Auxin-induced Degradation Experiments

Auxin treatment was performed by transferring young adult worms (aged 12-14 hours post-L4) to bacteria-seeded plates containing auxin or 99% ethanol at specific time points, except for experiments in which L4s were transferred directly to bacteria-seeded plates containing auxin or ethanol (the 36 hour and 48 hour time points in Figures 3C and E and the experiments in Figures 5C and D). The natural auxin indole3-acetic acid (IAA) was purchased from Alfa Aesar (#A10556). A 400 mM stock solution in ethanol was prepared and was stored at 4°C for up to one month. Auxin was diluted to 100mM, and 100ul was spread onto NGM plates. Plates were allowed to dry before seeding with fresh OP50 culture. Plates were left at 20°C for 2-3 days in the dark to allow for bacterial lawn growth.

Imaging and Quantification

All images were acquired using a DeltaVision Personal DV system (Applied Precision) equipped with a 100X N.A. 1.40 oil-immersion objective (Olympus), resulting in an effective XY pixel spacing of 0.064 or 0.040 µm. Three-dimensional image stacks were collected at 0.2-µm Z spacing and processed by constrained, iterative deconvolution. Image scaling, analysis and maximum-intensity projections were performed using functions in the softWoRx software package.

For analysis of GFP::MSH-5 foci, sum projections were generated using ImageJ for each image of the germline. ImageJ plugins Cell Counter, ROI Manager, and Find Maxima were used to identify and quantify GFP::MSH-5 foci by row from transition zone to the end of pachytene. The threshold value was set depending on background conditions to ensure minimal signals were identified. Foci were only quantified if they co-localized with DAPI staining. For all genotypes, three germlines per genotype were analyzed and representative germlines are shown.

Graphing and Statistical Analysis

Data was analyzed using Python 3.8 and Prism for statistical significance. All datasets were tested for normality using the Shapiro-Wilk test. For Figures 1, 2 and 5B, Fisher’s exact test was used to determine significance. For Figures 3, 4, 5D and 6, Mann-Whitney U test was used to determine significance.

Author contributions

B.P., M.G., and N.B. designed the experiments. B.P., M.G., A.H., E.L., and V.O. performed the experiments. B.P. and N.B. analyzed the data. B.P. and N.B. wrote the initial draft of the manuscript and B.P., M.G., A.H. and N.B. revised the manuscript. N.B. acquired funding.

eLife Assessment

This is an important study examining the role of conserved PCH-2 protein at different stages of C. elegans meiosis. The authors use elegant molecular genetic approaches to provide convincing evidence to support their claims. The work will be of interest to scientists studying meiosis, DNA recombination, and chromosome segregation.

Reviewer #1 (Public review):

The conserved AAA-ATPase PCH-2 has been shown in several organisms including C. elegans to remodel classes of HORMAD proteins that act in meiotic pairing and recombination. In some organisms the impact of PCH-2 mutations is subtle but becomes more apparent when other aspects of recombination are perturbed. Patel et al. performed a set of elegant experiments in C. elegans aimed at identifying conserved functions of PCH-2. Their work provides such an opportunity because in C. elegans meiotically expressed HORMADs localize to meiotic chromosomes independently of PCH-2. Work in C. elegans also allows the authors to focus on nuclear PCH-2 functions as opposed to cytoplasmic functions also seen for PCH-2 in other organisms.

The authors performed the following experiments:

(1) They constructed C. elegans animals with SNPs that enabled them to measure crossing over in intervals that cover most of four of the six chromosomes. They then showed that double-crossovers, which were common on most of the four chromosomes in wild-type, were absent in pch-2. They also noted shifts in crossover distribution in the four chromosomes.

(2) Based on the crossover analysis and previous studies they hypothesized that PCH-2 plays a role at an early stage in meiotic prophase to regulate how SPO-11 induced double-strand breaks are utilized to form crossovers. They tested their hypothesis by performing ionizing irradiation and depleting SPO-11 at different stages in meiotic prophase in wild-type and pch-2 mutant animals. The authors observed that irradiation of meiotic nuclei in zygotene resulted in pch-2 nuclei having a larger number of nuclei with 6 or greater crossovers (as measured by COSA-1 foci) compared to wildtype. Consistent with this observation, SPO11 depletion, starting roughly in zygotene, also resulted in pch-2 nuclei having an increase in 6 or more COSA-1 foci compared to wild type. The increased number at this time point appeared beneficial because a significant decrease in univalents was observed.

(3) They then asked if the above phenotypes correlated with the localization of MSH-5, a factor that stabilizes crossover-specific DNA recombination intermediates. They observed that pch-2 mutants displayed an increase in MSH-5 foci at early times in meiotic prophase and an unexpectedly higher number at later times. They conclude based on the differences in early MSH-5 localization and the SPO-11 and irradiation studies that PCH-2 prevents early DSBs from becoming crossovers and early loading of MSH-5. By analyzing different HORMAD proteins that are defective in forming the closed conformation acted upon by PCH-2, they present evidence that MSH-5 loading was regulated by the HIM-3 HORMAD.

(4) They performed a crossover homeostasis experiment in which DSB levels were reduced. The goal of this experiment was to test if PCH-2 acts in crossover assurance. Interestingly, in this background PCH-2 negative nuclei displayed higher levels of COSA-1 foci compared to PCH-2 positive nuclei. This observation and a further test of the model suggested that "PCH-2's presence on the SC prevents crossover designation."

(5) Based on their observations indicating that early DSBS are prevented from becoming crossovers by PCH-2, the authors hypothesized that the DNA damage kinase CHK-2 and PCH-2 act to control how DSBs enter the crossover pathway. This hypothesis was developed based on their finding that PCH-2 prevents early DSBs from becoming crossovers and previous work showing that CHK-2 activity is modulated during meiotic recombination progression. They tested their hypothesis using a mutant synaptonemal complex component that maintains high CHK-2 activity that cannot be turned off to enable crossover designation. Their finding that the pch-2 mutation suppressed the crossover defect (as measured by COSA-1 foci) supports their hypothesis.

Based on these studies the authors provide convincing evidence that PCH-2 prevents early DSBs from becoming crossovers and controls the number and distribution of crossovers to promote a regulated mechanism that ensures the formation of obligate crossovers and crossover homeostasis. As the authors note, such a mechanism is consistent with earlier studies suggesting that early DSBs could serve as "scouts" to facilitate homolog pairing or to coordinate the DNA damage response with repair events that lead to crossing over. The detailed mechanistic insights provided in this work will certainly be used to better understand functions for PCH-2 in meiosis in other organisms. My comments below are aimed at improving the clarity of the manuscript.

Comments

(1) It appears from reading the Materials and Methods that the SNPs used to measure crossing over were obtained by mating Hawaiian and Bristol strains. It is not clear to this reviewer how the SNPs were introduced into the animals. Was crossing over measured in a single animal line? Were the wild-type and pch-2 mutations made in backgrounds that were isogenic with respect to each other? This is a concern because it is not clear, at least to this reviewer, how much of an impact crossing different ecotypes will have on the frequency and distribution of recombination events (and possibly the recombination intermediates that were studied).

(2) The authors state that in pch-2 mutants there was a striking shift of crossovers (line 135) to the PC end for all of the four chromosomes that were tested. I looked at Figure 1 for some time and felt that the results were more ambiguous. Map distances seemed similar at the PC end for wildtype and pch-2 on Chrom. I. While the decrease in crossing over in pch-2 appeared significant for Chrom. I and III, the results for Chrom. IV, and Chrom. X. seemed less clear. Were map distances compared statistically? At least for this reviewer the effects on specific intervals appear less clear and without a bit more detail on how the animals were constructed it's hard for me to follow these conclusions.

(3) Figure 2. I'm curious why non-irradiated controls were not tested side-by-side for COSA-1 staining. It just seems like a nice control that would strengthen the authors' arguments.

(4) Figure 3. It took me a while to follow the connection between the COSA-1 staining and DAPI staining panels (12 hrs later). Perhaps an arrow that connects each set of time points between the panels or just a single title on the X-axis that links the two would make things clearer.

Reviewer #2 (Public review):

Summary:

This paper has some intriguing data regarding the different potential roles of Pch-2 in ensuring crossing over. In particular, the alterations in crossover distribution and Msh-5 foci are compelling. My main issue is that some of the models are confusingly presented and would benefit from some reframing. The role of Pch-2 across organisms has been difficult to determine, the ability to separate pairing and synapsis roles in worms provides a great advantage for this paper.

Strengths:

Beautiful genetic data, clearly made figures. Great system for studying the role of Pch-2 in crossing over.

Weaknesses:

(1) For a general audience, definitions of crossover assurance, crossover eligible intermediates, and crossover designation would be helpful. This applies to both the proposed molecular model and the cytological manifestation that is being scored specifically in C. Elegans.

(2) Line 62: Is there evidence that DSBs are introduced gradually throughout the early prophase? Please provide references.

(3) Do double crossovers show strong interference in worms? Given that the PC is at the ends of chromosomes don't you expect double crossovers to be near the chromosome ends and thus the PC?

(4) Line 155 - if the previous data in Deshong et al is helpful it would be useful to briefly describe it and how the experimental caveats led to misinterpretation (or state that further investigation suggests a different model etc.). Many readers are unlikely to look up the paper to find out what this means.

(5) Line 248: I am confused by the meaning of crossover assurance here - you see no difference in the average number of COSA-1 foci in Pch-2 vs. wt at any time point. Is it the increase in cells with >6 COSA-1 foci that shows a loss of crossover assurance? That is the only thing that shows a significant difference (at the one time point) in COSA-1 foci. The number of dapi bodies shows the loss of Pch-2 increases crossover assurance (fewer cells with unattached homologs). So this part is confusing to me. How does reliably detecting foci vs. DAPI bodies explain this?

(6) Line 384: I am confused. I understand that in the dsb-2/pch2 mutant there are fewer COSA-1 foci. So fewer crossovers are designated when DSBs are reduced in the absence of PCH-2. How then does this suggest that PCH-2's presence on the SC prevents crossover designation? Its absence is preventing crossover designation at least in the dsb-2 mutant.

(7) Discussion Line 535: How do you know that the crossovers that form near the PCs are Class II and not the other way around? Perhaps early forming Class I crossovers give time for a second Class II crossover to form. In budding yeast, it is thought that synapsis initiation sites are likely sites of crossover designation and class I crossing over. Also, the precursors that form class I and II crossovers may be the same or highly similar to each other, such that Pch-2's actions could equally affect both pathways.

Reviewer #3 (Public review):

Summary:

This manuscript describes an in-depth analysis of the effect of the AAA+ ATPase PCH-2 on meiotic crossover formation in C. elegant. The authors reach several conclusions, and attempt to synthesize a 'universal' framework for the role of this factor in eukaryotic meiosis.

Strengths:

The manuscript makes use of the advantages of the 'conveyor' belt system within the c.elegans reproductive tract, to enable a series of elegant genetic experiments.

Weaknesses:

A weakness of this manuscript is that it heavily relies on certain genetic/cell biological assays that can report on distinct crossover outcomes, without clear and directed control over other aspects and variables that might also impact the final repair outcome. Such assays are currently out of reach in this model system.

In general, this manuscript could be more generally accessible to non-C.elegans readers. Currently, the manuscript is hard to digest for non-experts (even if meiosis researchers). In addition, the authors should be careful to consider alternative explanations for certain results. At several steps in the manuscript, results could ostensibly be caused by underlying defects that are currently unknown (for example, can we know for sure that pch-2 mutants do not suffer from altered DSB patterning, and how can we know what the exact functional and genetic interactions between pch-2 and HORMAD mutants tell us?). Alternative explanations are possible and it would serve the reader well to explicitly name and explain these options throughout the manuscript.

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1 (Public review):

The conserved AAA-ATPase PCH-2 has been shown in several organisms including C. elegans to remodel classes of HORMAD proteins that act in meiotic pairing and recombination. In some organisms the impact of PCH-2 mutations is subtle but becomes more apparent when other aspects of recombination are perturbed. Patel et al. performed a set of elegant experiments in C. elegans aimed at identifying conserved functions of PCH-2. Their work provides such an opportunity because in C. elegans meiotically expressed HORMADs localize to meiotic chromosomes independently of PCH-2. Work in C. elegans also allows the authors to focus on nuclear PCH-2 functions as opposed to cytoplasmic functions also seen for PCH-2 in other organisms.

The authors performed the following experiments:

(1) They constructed C. elegans animals with SNPs that enabled them to measure crossing over in intervals that cover most of four of the six chromosomes. They then showed that doublecrossovers, which were common on most of the four chromosomes in wild-type, were absent in pch-2. They also noted shifts in crossover distribution in the four chromosomes.

(2) Based on the crossover analysis and previous studies they hypothesized that PCH-2 plays a role at an early stage in meiotic prophase to regulate how SPO-11 induced double-strand breaks are utilized to form crossovers. They tested their hypothesis by performing ionizing irradiation and depleting SPO-11 at different stages in meiotic prophase in wild-type and pch-2 mutant animals. The authors observed that irradiation of meiotic nuclei in zygotene resulted in pch-2 nuclei having a larger number of nuclei with 6 or greater crossovers (as measured by COSA-1 foci) compared to wildtype. Consistent with this observation, SPO11 depletion, starting roughly in zygotene, also resulted in pch-2 nuclei having an increase in 6 or more COSA-1 foci compared to wild type. The increased number at this time point appeared beneficial because a significant decrease in univalents was observed.

(3) They then asked if the above phenotypes correlated with the localization of MSH-5, a factor that stabilizes crossover-specific DNA recombination intermediates. They observed that pch-2 mutants displayed an increase in MSH-5 foci at early times in meiotic prophase and an unexpectedly higher number at later times. They conclude based on the differences in early MSH-5 localization and the SPO-11 and irradiation studies that PCH-2 prevents early DSBs from becoming crossovers and early loading of MSH-5. By analyzing different HORMAD proteins that are defective in forming the closed conformation acted upon by PCH-2, they present evidence that MSH-5 loading was regulated by the HIM-3 HORMAD.

(4) They performed a crossover homeostasis experiment in which DSB levels were reduced. The goal of this experiment was to test if PCH-2 acts in crossover assurance. Interestingly, in this background PCH-2 negative nuclei displayed higher levels of COSA-1 foci compared to PCH-2 positive nuclei. This observation and a further test of the model suggested that "PCH-2's presence on the SC prevents crossover designation."

(5) Based on their observations indicating that early DSBS are prevented from becoming crossovers by PCH-2, the authors hypothesized that the DNA damage kinase CHK-2 and PCH2 act to control how DSBs enter the crossover pathway. This hypothesis was developed based on their finding that PCH-2 prevents early DSBs from becoming crossovers and previous work showing that CHK-2 activity is modulated during meiotic recombination progression. They tested their hypothesis using a mutant synaptonemal complex component that maintains high CHK-2 activity that cannot be turned off to enable crossover designation. Their finding that the pch-2 mutation suppressed the crossover defect (as measured by COSA-1 foci) supports their hypothesis.

Based on these studies the authors provide convincing evidence that PCH-2 prevents early DSBs from becoming crossovers and controls the number and distribution of crossovers to promote a regulated mechanism that ensures the formation of obligate crossovers and crossover homeostasis. As the authors note, such a mechanism is consistent with earlier studies suggesting that early DSBs could serve as "scouts" to facilitate homolog pairing or to coordinate the DNA damage response with repair events that lead to crossing over. The detailed mechanistic insights provided in this work will certainly be used to better understand functions for PCH-2 in meiosis in other organisms. My comments below are aimed at improving the clarity of the manuscript.

We thank the reviewer for their concise summary of our manuscript and their assessment of our work as “convincing” and providing “detailed mechanistic insight.”

Comments

(1) It appears from reading the Materials and Methods that the SNPs used to measure crossing over were obtained by mating Hawaiian and Bristol strains. It is not clear to this reviewer how the SNPs were introduced into the animals. Was crossing over measured in a single animal line? Were the wild-type and pch-2 mutations made in backgrounds that were isogenic with respect to each other? This is a concern because it is not clear, at least to this reviewer, how much of an impact crossing different ecotypes will have on the frequency and distribution of recombination events (and possibly the recombination intermediates that were studied).

We have clarified these issues in the Materials and Methods of our updated preprint. The control and pch-2 mutants were isogenic in either the Bristol or Hawaiian backgrounds. Control lines were the original Bristol and Hawaiian lines and pch-2 mutants were originally made in the Bristol line and backcrossed at least 3 times before analysis. Hawaiian pch-2 mutants were made by backcrossing pch-2 mutants at least 8 times to the Hawaiian background and verifying the presence of Hawaiian SNPs on all chromosomes tested in the recombination assay. To perform the recombination assays, these lines were crossed to generate the relevant F1s.

(2) The authors state that in pch-2 mutants there was a striking shift of crossovers (line 135) to the PC end for all of the four chromosomes that were tested. I looked at Figure 1 for some time and felt that the results were more ambiguous. Map distances seemed similar at the PC end for wildtype and pch-2 on Chrom. I. While the decrease in crossing over in pch-2 appeared significant for Chrom. I and III, the results for Chrom. IV, and Chrom. X. seemed less clear. Were map distances compared statistically? At least for this reviewer the effects on specific intervals appear less clear and without a bit more detail on how the animals were constructed it's hard for me to follow these conclusions.

We hope that the added details above makes the results of these assays more clear. Map distances were compared and did not satisfy statistical significance, except where indicated. While we agree that the comparisons between control animals and pch-2 mutants may seem less clear with individual chromosomes, we argue that more general, consistent patterns become clear when analyzing multiple chromosomes. Indeed, this is why we expanded our recombination analysis beyond Chromosome III and the X Chromosome, as reported in Deshong, 2014. We have edited this sentence to: “Moreover, there was a striking and consistent shift of crossovers to the PC end of all four chromosomes tested.”

(3) Figure 2. I'm curious why non-irradiated controls were not tested side-by-side for COSA-1 staining. It just seems like a nice control that would strengthen the authors' arguments.

We have added these controls in the updated preprint as Figure 2B.

(4) Figure 3. It took me a while to follow the connection between the COSA-1 staining and DAPI staining panels (12 hrs later). Perhaps an arrow that connects each set of time points between the panels or just a single title on the X-axis that links the two would make things clearer.

To make this figure more clear, we have generated two different cartoons for the assay that scores GFP::COSA-1 foci and the assay that scores bivalents. We have also edited this section of the results to make it more clear.

Reviewer #2 (Public review):

Summary:

This paper has some intriguing data regarding the different potential roles of Pch-2 in ensuring crossing over. In particular, the alterations in crossover distribution and Msh-5 foci are compelling. My main issue is that some of the models are confusingly presented and would benefit from some reframing. The role of Pch-2 across organisms has been difficult to determine, the ability to separate pairing and synapsis roles in worms provides a great advantage for this paper.

Strengths:

Beautiful genetic data, clearly made figures. Great system for studying the role of Pch-2 in crossing over.

We thank the reviewers for their constructive and useful summary of our manuscript and the analysis of its strengths.

Weaknesses:

(1) For a general audience, definitions of crossover assurance, crossover eligible intermediates, and crossover designation would be helpful. This applies to both the proposed molecular model and the cytological manifestation that is being scored specifically in C. elegans.

We have made these changes in an updated preprint.

(2) Line 62: Is there evidence that DSBs are introduced gradually throughout the early prophase? Please provide references.

We have referenced Woglar and Villeneuve 2018 and Joshi et. al. 2015 to support this statement in the updated preprint.

(3) Do double crossovers show strong interference in worms? Given that the PC is at the ends of chromosomes don't you expect double crossovers to be near the chromosome ends and thus the PC?

Despite their rarity, double crossovers do show interference in worms. However, the PC is limited to one end of the chromosome. Therefore, even if interference ensures the spacing of these double crossovers, the preponderance of one of these crossovers toward one end (and not both ends) suggest something functionally unique about the PC end.

(4) Line 155 - if the previous data in Deshong et al is helpful it would be useful to briefly describe it and how the experimental caveats led to misinterpretation (or state that further investigation suggests a different model etc.). Many readers are unlikely to look up the paper to find out what this means.

We have added this to the updated preprint: “We had previously observed that meiotic nuclei in early prophase were more likely to produce crossovers when DSBs were induced by the Mos transposon in pch-2 mutants than in control animals but experimental caveats limited our ability to properly interpret this experiment.”

(5) Line 248: I am confused by the meaning of crossover assurance here - you see no difference in the average number of COSA-1 foci in Pch-2 vs. wt at any time point. Is it the increase in cells with >6 COSA-1 foci that shows a loss of crossover assurance? That is the only thing that shows a significant difference (at the one time point) in COSA-1 foci. The number of dapi bodies shows the loss of Pch-2 increases crossover assurance (fewer cells with unattached homologs). So this part is confusing to me. How does reliably detecting foci vs. DAPI bodies explain this?

We have removed this section to avoid confusion.

(6) Line 384: I am confused. I understand that in the dsb-2/pch2 mutant there are fewer COSA-1 foci. So fewer crossovers are designated when DSBs are reduced in the absence of PCH-2.

How then does this suggest that PCH-2's presence on the SC prevents crossover designation? Its absence is preventing crossover designation at least in the dsb-2 mutant.

We have tried to make this more clear in the updated preprint. In this experiment, we had identified three possible explanations for why PCH-2 persists on some nuclei that do not have GFP::COSA-1 foci: 1) PCH-2 removal is coincident with crossover designation; 2) PCH-2 removal depends on crossover designation; and 3) PCH-2 removal facilitates crossover designation. The decrease in the number of GFP::COSA-1 foci in dsb2::AID;pch-2 mutants argues against the first two possibilities, suggesting that the third might be correct. We have edited the sentence to read: “These data argue against the possibility that PCH-2’s removal from the SC is simply in response to or coincident with crossover designation and instead, suggest that PCH-2’s removal from the SC somehow facilitates crossover designation and assurance.”

(7) Discussion Line 535: How do you know that the crossovers that form near the PCs are Class II and not the other way around? Perhaps early forming Class I crossovers give time for a second Class II crossover to form. In budding yeast, it is thought that synapsis initiation sites are likely sites of crossover designation and class I crossing over. Also, the precursors that form class I and II crossovers may be the same or highly similar to each other, such that Pch-2's actions could equally affect both pathways.

We do not know that the crossovers that form near the PC are Class II but hypothesize that they are based on the close, functional relationship that exists between Class I crossovers and synapsis and the apparent antagonistic relationship that exists between Class II crossovers and synapsis. We agree that Class I and Class II crossover precursors are likely to be the same or highly similar, exhibit extensive crosstalk that may complicate straightforward analysis and PCH-2 is likely to affect both, as strongly suggested by our GFP::MSH-5 analysis. We present this hypothesis based on the apparent relationship between PCH-2 and synapsis in several systems but agree that it needs to be formally tested. We have tried to make this argument more clear in the updated preprint.

Reviewer #3 (Public review):

Summary:

This manuscript describes an in-depth analysis of the effect of the AAA+ ATPase PCH-2 on meiotic crossover formation in C. elegant. The authors reach several conclusions, and attempt to synthesize a 'universal' framework for the role of this factor in eukaryotic meiosis.

Strengths:

The manuscript makes use of the advantages of the 'conveyor' belt system within the c.elegans reproductive tract, to enable a series of elegant genetic experiments.

We thank this reviewer for the useful assessment of our manuscript and the articulation of its strengths.

Weaknesses:

A weakness of this manuscript is that it heavily relies on certain genetic/cell biological assays that can report on distinct crossover outcomes, without clear and directed control over other aspects and variables that might also impact the final repair outcome. Such assays are currently out of reach in this model system.

In general, this manuscript could be more generally accessible to non-C.elegans readers. Currently, the manuscript is hard to digest for non-experts (even if meiosis researchers). In addition, the authors should be careful to consider alternative explanations for certain results. At several steps in the manuscript, results could ostensibly be caused by underlying defects that are currently unknown (for example, can we know for sure that pch-2 mutants do not suffer from altered DSB patterning, and how can we know what the exact functional and genetic interactions between pch-2 and HORMAD mutants tell us?). Alternative explanations are possible and it would serve the reader well to explicitly name and explain these options throughout the manuscript.

We have made the manuscript more accessible to non-C. elegans readers and discuss alternate explanations for specific results in the updated preprint.

Recommendations for the authors:

Reviewing Editor Comments:

(1) Please provide 'n' values for each experiment.

n values are now included in the Figure legends for each experiment.

(2) Line 129: Please represent the DCOs as percent or fraction (1%-9.8%, instead of 1-13).

We have made this change.

(3) Figure 3A legend: the grey bar should read 20hr. COSA-1/ 32 hr DAPI. In Figure 3E, it is not clear why 36hr Auxin and 34hr Auxin show a significant difference in DAPI bodies between control and pch-2, but 32hr Auxin treatment does not. Here again 'n' values will help.

We have made this change. We also are not sure why the 32 hour auxin treatment did not show a significant difference in DAPI stained bodies. We have included the n values, which are not very different between timepoints and therefore are unlikely to explain the difference. The difference may reflect the time that it takes for SPO-11 function to be completely abrogated.

(4) Line 360: Please provide the fraction of PCH-2 positive nuclei in dsb-2.

We have made this change.

Please also address all reviewer comments.

Reviewer #1 (Recommendations for the authors):

(1) Page 3, line 52. While I agree that crossing over is important to generate new haplotypes, work has suggested that the contribution by an independent assortment of homologs to generate new haplotypes is likely to be significantly greater. One reference for this is: Veller et al. PNAS 116:1659.

We deeply appreciate this reviewer pointing us to this paper, especially since it argues that controlling crossover distribution contributes to gene shuffling and now cite it in our introduction! While we agree that this paper concludes that independent assortment likely explains the generation of new haplotypes to a greater degree than crossovers, the authors performed this analysis with human chromosomes and explicitly include the caveat that their modeling assumes uniform gene density across chromosomes. For example, we know this is not true in C. elegans. It would be interesting to perform the same analysis with C. elegans chromosomes in control and pch-2 mutants, taking into account this important difference.

(2) Figure 2. It would really help the reader if an arrow and text were shown below each irradiation sign to indicate the stage in meiosis in which the irradiation was done as well as another arrow in the late pachytene box to show when the COSA-1 foci were analyzed. In general, having text in the figures that help stage the timing in meiosis would help the non C. elegans reader. This is also an issue where staging of C. elegans is shown (Figure 4).

We have made these changes to Figure 2. To help readers interpret Figure 4, we have added TZ and LP to the graphs in Figure 4B and 4D and indicated what these acronyms (transition zone and late pachytene, respectively) are in the Figure legend.

(3) Page 12, line 288. It would be valuable to first outline why the him3-R93Y and htp-3H96Y alleles were chosen. This was eventually done on Page 13, but introducing this earlier would help the reader.

We have introduced these mutations earlier in the manuscript.

(4) Page 13, line 323. A one sentence description of the OLLAS tagging system would be useful.

We have added this sentence: “we generated wildtype animals and pch-2 mutants with both GFP::MSH-5 and a version of COSA-1 that has been endogenously tagged at the Nterminus with the epitope tag, OLLAS, a fusion of the E. coli OmpF protein and the mouse Langerin extracellular domain”

Reviewer #2 (Recommendations for the authors):

(1) The title is a little awkward. Consider: PCH-2 controls the number and distribution of crossovers in C. elegans by antagonizing their formation

We have made this change.

(2) Abstract:

Consider removing "that is observed" from line 20.

We have made this change.

I'm confused by the meaning of "reinforcement of crossover-eligible intermediates" from line 27.

We have removed this phrase from the abstract.

A definition of crossover assurance would be helpful in the abstract.

We have added this to the abstract: “This requirement is known as crossover assurance and is one example of crossover control.”

(3) Line 36: I know a stickler but many meioses only produce one haploid gamete (mammalian oocytes, for example)

Thanks for the reminder! We have removed the “four” from this sentence.

(4) Line 284 - are you defining MSH-5 foci as crossover-eligible intermediates? If so, please state this earlier.

We have added this to the introduction to this section of the results: “In C. elegans, these crossover-eligible intermediates can be visualized by the loading of the pro-crossover factor MSH-5, a component of the meiosis-specific MutSγ complex that stabilizes crossover-specific DNA repair intermediates called joint molecules”

(5) Can the control be included in Figure S1?

We have made this change.

(6) Can you define that crossover designation is the formation of a COSA-1 focus?

We did this in the section introducing GFP::MSH-5: “In the spatiotemporally organized meiotic nuclei of the germline, a functional GFP tagged version of MSH-5, GFP::MSH-5, begins to form a few foci in leptotene/zygotene (the transition zone), becoming more numerous in early pachytene before decreasing in number in mid pachytene to ultimately colocalize with COSA-1 marked sites in late pachytene in a process called designation”

(7) Would it be easier to see the effect of DSB to crossover eligible intermediates in Spo-11, Pch-2 vs. Spo-11 mutant with irradiation using your genetic maps? At least for early vs. late breaks?

Unfortunately, irradiation does not show the same bias towards genomic location that endogenous double strand breaks do so it is unlikely to recapitulate the effects on the genetic map.

Version 2

Abstract

Abstract

Meiotic crossover recombination is essential for both accurate chromosome segregation and the generation of new haplotypes for natural selection to act upon. This requirement is known as crossover assurance and is one example of crossover control. While the conserved role of the ATPase, PCH-2, during meiotic prophase has been enigmatic, a universal phenotype when pch-2 or its orthologs are mutated is a change in the number and distribution of meiotic crossovers. Here, we show that PCH-2 controls the number and distribution of crossovers by antagonizing their formation. This antagonism produces different effects at different stages of meiotic prophase: early in meiotic prophase, PCH-2 prevents double strand breaks from becoming crossover-eligible intermediates, limiting crossover formation at sites of initial double strand break formation and homolog interactions. Later in meiotic prophase, PCH-2 winnows the number of crossover-eligible intermediates, contributing to the designation of crossovers and ultimately, crossover assurance. We also demonstrate that PCH-2 accomplishes this regulation through the meiotic HORMAD, HIM-3. Our data strongly support a model in which PCH-2’s conserved role is to remodel meiotic HORMADs throughout meiotic prophase to destabilize crossover-eligible precursors, coordinate meiotic recombination with synapsis, and contribute to the progressive implementation of meiotic recombination, guaranteeing crossover control.

Introduction

Meiosis is a specialized type of cell division that reduces chromosome number by half, resulting in the production of genetically diverse haploid gametes, so that fertilization during sexual reproduction restores diploidy. This process occurs in two stages: meiosis I, in which homologous chromosomes are partitioned, and meiosis II, in which sister chromatid are segregated. The regulation of meiosis is crucial for ensuring that both genetic recombination and chromosome segregation occur accurately. Errors in human meiosis are associated with birth defects, such as Down and Turner syndromes, infertility and miscarriages, underscoring the importance of understanding meiosis to human health (Hassold and Hunt, 2001).

During prophase I, homologous chromosomes pair and this pairing is stabilized through a process called synapsis, in which a protein structure called the synaptonemal complex (SC) holds homologs together. This close association between homologs during synapsis facilitates crossover formation, the process in which double strand breaks (DSBs) are deliberately introduced into the genome, repaired by meiosis-specific mechanisms that exchange DNA between homologous chromosomes and generate the chiasmata, or linkage, that promotes accurate meiotic chromosome segregation. Therefore, disruptions in pairing, synapsis, or recombination prevent the formation of chiasmata and can lead to meiotic errors such as nondisjunction, resulting in aneuploid gametes.

In addition to the fundamental role that recombination plays in ensuring accurate chromosome segregation, crossover recombination accomplishes another important function: it generates new haplotypes for natural selection to act upon to drive evolution. Thus, to assure a random assortment of alleles on a population level, the distribution of crossovers may be as tightly regulated as their number (Veller et al., 2019). The significance of controlling both crossover number and distribution is clearly illustrated by the existence of mechanisms such as crossover assurance, in which every pair of homologous chromosomes gets at least one crossover; crossover homeostasis, in which the number of crossovers remains relatively invariant even if the number of recombination precursors change; and crossover interference, in which the presence of a crossover inhibits the formation of a crossover nearby (Gray and Cohen, 2016). DSBs typically vastly outnumber crossovers in most organisms and are introduced gradually throughout early prophase (Joshi et al., 2015; Woglar and Villeneuve, 2018). Therefore, to accomplish this precise level of control, meiotic crossover recombination and the decision about which DSBs become crossover-eligible intermediates, and eventually, which crossover-eligible intermediates get designated as crossovers, is implemented progressively throughout meiotic prophase (Cole et al., 2012; Joshi et al., 2015; Morgan et al., 2021; Yokoo et al., 2012). In many systems, transitions from DSBs to crossover-eligible intermediates, and crossover-eligible intermediates to crossovers, can be molecularly and/or cytologically monitored.

PCH-2, also known as TRIP13 in mammals, is an evolutionarily ancient AAA-ATPase that plays a significant role in regulating meiosis across different organisms, including M. musculus (mice), S. cerevisiae (budding yeast), D. melanogaster (fruit flies), and C. elegans (worms) (Bhalla, 2023). PCH-2 and its orthologs structurally remodel a family of proteins with HORMA domains (HORMADs) to control their function (Gu et al., 2022). HORMADs participate in a variety of signaling events and can exist in at least three structurally distinct conformations: a “closed” conformation, which they adopt when they bind a short peptide sequence in their own protein sequence or another protein (also called a closure motif) and their C-terminus wraps around this motif to stabilize the interaction; an “open” conformation when unbound, in which their C-terminus is discretely tucked against the HORMA domain; and an “extended” conformation, which is an intermediate between the two (Gu et al., 2022). PCH-2 and its orthologs convert the closed version of HORMADs to the open or extended versions, playing an important role in recycling HORMADs during signaling. During meiosis, closed versions of meiotic HORMADs assemble on chromosomes to form meiotic chromosome axes, which are essential for pairing, synapsis and recombination between homologous chromosomes (Kim et al., 2014). The remodeling of meiotic HORMADs by PCH-2 to an “open” or “extended” conformation is thought to reduce the levels of HORMADs on chromosomes (Borner et al., 2008; Cuacos et al., 2021; Lambing et al., 2015; Wojtasz et al., 2009) and/or increase their dynamic association and dissociation (Russo et al., 2023), modulating and coordinating homolog pairing, synapsis and recombination during prophase.

The conserved role of the PCH-2/HORMAD module in meiosis has been difficult to characterize, in part, we have argued, because of the evolutionary innovation that is an inherent aspect of sexual reproduction (Bhalla, 2023). Moreover, in some systems, such as budding yeast and plants (Herruzo et al., 2021; Yang et al., 2020), PCH-2 orthologs not only remodel meiotic HORMADs on meiotic chromosomes but also perform this function in the cytoplasm to make meiotic HORMADs available for their role(s) in meiotic nuclei. This dual role can complicate functional analyses, particularly in pch-2 null mutants. However, all meiotic systems exhibit defects in the number and distribution of crossovers when PCH-2 function is abrogated by mutation (Deshong et al., 2014; Joshi et al., 2009; Joyce and McKim, 2009; Lambing et al., 2015; Roig et al., 2010; Zanders and Alani, 2009). Unfortunately, the lack of a clear pattern when analyzing these defects in recombination in pch-2 mutants has contributed to an inability to develop a unified, integrated model of PCH-2 function in the field.

In C. elegans, meiotic HORMADs localize to meiotic chromosomes independently of PCH-2 (Deshong et al., 2014). Moreover, the progressive implementation of meiotic recombination can be cytologically monitored in meiotic nuclei that are organized both spatially and temporally in the C. elegans germline (Woglar and Villeneuve, 2018; Yokoo et al., 2012). Here, we exploit this system to show that PCH-2 is required to control the number and distribution of crossovers by antagonizing their formation. This antagonism produces different consequences depending on the stage of meiotic prophase. In early meiotic prophase, PCH-2 inhibits DSBs from becoming crossover-eligible intermediates, ensuring that crossovers are more widely distributed than sites of initial DSB formation and/or homolog interactions. Later in meiotic prophase, PCH-2 is responsible for winnowing the numbers of crossover-eligible intermediates on synapsed chromosomes, contributing to the designation of crossovers and ultimately, crossover assurance. Genetic analysis demonstrates that PCH-2’s regulation of crossover-eligible intermediates is through one of three essential meiotic HORMADs, HIM-3. Finally, we link PCH-2’s effect on early DSBs in early meiotic prophase to cell cycle stage, demonstrating that both limit early DSBs from becoming crossovers. We propose that PCH-2’s remodeling of HIM-3 on meiotic chromosomes destabilizes crossover-eligible intermediates throughout meiotic prophase, contributing to the progressive implementation of meiotic recombination to control the number and distribution of crossovers, also known as crossover control.

Results

PCH-2 controls the number and distribution of crossovers in similar patterns on multiple chromosomes

We had previously shown that loss of PCH-2 reduced the frequency of double crossovers and genetic length of both an autosome (chromosome III) and the X chromosome, albeit not uniformly among genetic intervals (Deshong et al., 2014). To determine whether a more obvious pattern could be observed, we expanded our analysis by monitoring recombination genetically in both wildtype and pch-2 mutant animals using five single nucleotide polymorphisms (SNPs) that spanned 95% of Chromosomes, I, III, IV and the X chromosome (Figure 1). We excluded Chromosome II from our analysis because of the potential difficulty combining Hawaiian SNPs with the pch-2 mutation, which is linked to Chromosome II, and Chromosome V because of our use of the bcIs39 transgene to identify cross progeny, which may disrupt recombination on that chromosome.

PCH-2 controls the number and distribution of crossovers in similar patterns on multiple chromosomes.

Genetic analysis of meiotic recombination in wildtype and pch-2 mutants. DCO indicates double crossovers. Physical and genetic maps of Chromosome I, III, IV and the X chromosome are depicted to scale. Genetic distance is shown in centimorgans. A * indicates a p-value < 0.05, a ** indicates a p value < 0.01 and a *** indicates p-value < 0.001.

In wildtype animals, we observed multiple double crossovers, ranging from 1-13, depending on the chromosome. In a majority of these double crossovers (69%), one crossover was at the end of the chromosome where pairing and synapsis initiate, also called the Pairing Center (PC) (MacQueen et al., 2005), suggesting some relationship between where chromosomes might make initial contacts and the likelihood of double crossovers. On all chromosomes analyzed, we observed no double crossovers in pch-2 mutants and this difference was statistically significant for Chromosomes I, III and X. Moreover, there was a striking and consistent shift of crossovers to the PC end of all four chromosomes tested. This shift in the distribution of crossovers to the PC ends of chromosomes was generally accompanied by a reduction in crossovers in the center of chromosomes. In C. elegans, the center of chromosomes are where double strand breaks are less numerous (Nadarajan et al., 2021; Yu et al., 2016) and where genes are more abundant. In the case of the X chromosome and chromosome III, recombination at the non-PC end was also reduced and appeared to more closely resemble the physical map at this end of these chromosomes. Thus, PCH-2 ensures a wider distribution of crossovers across chromosomes, away from regions that are more likely to undergo early homolog interactions (PC ends) and with more double strand breaks (both the PC and non-PC ends of chromosomes), and towards the center of chromosomes, where double strand breaks are less abundant.

PCH-2 prevents exogenous DSBs early in meiotic prophase from becoming crossovers

We previously showed that PCH-2 promotes crossover formation and crossover assurance through its regulation of the meiotic HORMAD, HIM-3, in C. elegans (Russo et al., 2023). Moreover, this promoting role seems linked to its localization to the synaptonemal complex (Patel et al., 2023). However, the observed shift in the distribution of crossovers in pch-2 mutants suggests that PCH-2 may also play a role in inhibiting crossovers and that this role may be occurring in early meiotic prophase, when chromosomes are undergoing initial homolog interactions. We had previously observed that meiotic nuclei in early prophase were more likely to produce crossovers when DSBs were induced by excision of the Mos transposon in pch-2 mutants than in control animals but experimental caveats limited our ability to properly interpret this experiment (Deshong et al., 2014).

To explicitly test this possibility, we took advantage of both the spatiotemporal organization of meiotic nuclei in the C. elegans germline, and the observation that nuclei travel in an assembly line process at a stereotypical pace to late pachytene (Jaramillo-Lambert et al., 2007), where we can cytologically assess crossover formation by staining for the essential crossover factor, COSA-1 (Yokoo et al., 2012). We performed irradiation experiments to introduce exogenous DSBs at different timepoints during meiotic prophase and analyzed PCH-2’s role in crossover formation (Figure 2A).

PCH-2 prevents exogenous double strand breaks from becoming crossovers in early meiotic prophase.

A. Illustration of the irradiation experiments in control and pch-2 mutants. Box indicates late pachytene, the area where GFP::COSA-1 foci are analyzed. B. Fraction of meiotic nuclei with less than 6, 6, or greater than 6 GFP::COSA-1 foci in control animals (yellow, n = 446) and pch-2 mutants (blue, n = 552). C. Meiotic nuclei in control animals and pch-2 mutants 8 hours post irradiation stained for DAPI (magenta) and GFP::COSA-1 (green). Scale bar is 4 um. D. Fraction of meiotic nuclei with less than 6, 6, or greater than 6 GFP::COSA-1 foci in control animals (yellow, n = 143) and pch-2 mutants (blue, n = 125) 8 hours post irradiation. E. Meiotic nuclei in control animals and pch-2 mutants 24 hours post irradiation with DAPI (magenta) and GFP::COSA-1 (green). F. Fraction of meiotic nuclei with less than 6, 6, or greater than 6 GFP::COSA-1 foci in control animals (n = 179) and pch-2 mutants (n = 378) 24 hours post irradiation. A *** indicates p-value < 0.001, and a **** indicates p-value < 0.0001.

In this experiment, we tested the effects of ionizing radiation at two distinct time points: 8 hours after irradiation, when we can monitor crossover formation in late pachytene nuclei that received exogenous DSBs in mid-pachytene, and 24 hours after irradiation, when we monitor crossover formation in late pachytene nuclei that received exogenous DSBs in leptotene/zygotene, also known as the transition zone in C. elegans (Figure 2A). In order to detect crossovers in late pachytene, control and pch-2 mutants germlines were stained for GFP::COSA-1, which allowed us to visualize crossovers, and DAPI, which allowed us to visualize DNA. In unirradiated control worms, 94% of late pachytene nuclei have 6 GFP::COSA-1 foci, one per chromosome pair (Figure 2B). In unirradiated pch-2 mutants, the percentage of nuclei have 6 GFP::COSA-1 foci drops to 82%, primarily because of the significant increase in meiotic nuclei with less than 6 GFP::COSA-1 foci (17%) (Figure 2B), consistent with our previous report that loss of PCH-2 leads to a decrease in crossover formation (Deshong et al., 2014). At 8 hours after irradiation, pch-2 mutants’ defect in crossover assurance becomes even more pronounced. We observed a significant increase in nuclei containing less than 6 COSA-1 nuclei in pch-2 mutants (26%), when compared to control (9%) germlines (Figures 2C and 2D). However, we did not observe a significant number of nuclei with greater than 6 COSA-1 foci in either backgrounds. This observation reinforces previous findings that exogenous DSBs introduced in mid-pachytene do not affect the number of crossovers in wildtype animals (Yokoo et al., 2012).

Next, we examined crossover formation 24 hours post-irradiation (Figures 2E and 2F). For this time point, we also detected a significant loss of crossover assurance in pch-2 mutants, compared to control worms. However, in contrast to the 8 hour time point, we also saw a significant increase in nuclei containing less than 6 COSA-1 foci in both control (13%) and pch-2 mutants (26%), compared to unirradiated worms (p values = 0.0008 and < 0.0001, respectively, Fischer’s Exact Test), indicating that even in control animals, extra DSBs in early meiotic prophase can disrupt crossover assurance. In addition, we also observed a significant increase in nuclei with more than 6 COSA-1 foci in pch-2 mutants (wildtype: 2%, pch-2: 13%), demonstrating that loss of PCH-2 leads to increased crossover formation when nuclei in the transition zone get more DSBs. In other words, PCH-2 inhibits exogenous DSBs introduced in early meiotic prophase from becoming crossovers, supporting our recombination analysis (Figure 1).

PCH-2 prevents SPO-11-induced DSBs from becoming crossovers in early meiotic prophase

Our previous experiment indicated that PCH-2 is preventing exogenous DSBs in early meiotic prophase from becoming crossovers (Figure 2). To test whether this was also the case for programmed meiotic DSB formation, we used the auxin-inducible degradation (AID) system to remove the enzyme that is responsible for programmed meiotic DSBs, SPO-11 (spo-11::AID::3XFLAG) (Dernburg et al., 1998; Keeney et al., 1997). When we treated both control and pch-2 mutant worms for 36 hours with auxin (Figure 3A), we observed a complete loss of GFP::COSA-1 foci (Figure 3C). When these animals were treated with auxin for 48 hours, to allow nuclei without SPO-11 an additional 12 hours to reach diakinesis (Figure 3D), we observed 12 DAPI stained bodies, or univalents, which are the 6 homolog pairs that have not formed chiasmata (Figure 3F). Thus, we can reliably remove SPO-11 from meiotic nuclei throughout the germline and prevent crossover formation.

PCH-2 prevents SPO-11-induced double strand breaks from becoming crossovers in early meiotic prophase.

A. Illustration of the SPO-11 depletion experiment to assay GFP::COSA-1 in control animals and pch-2 mutants at different timepoints of auxin treatment. Each timepoint indicates when SPO-11 is depleted in the germline with auxin induced degradation. Box indicates late pachytene, the area where GFP::COSA-1 foci are analyzed. B. Representative images of meiotic nuclei in control animals and pch-2 mutants treated with auxin for 20 hours, stained for DAPI (magenta) and GFP::COSA-1 (green). Scale bar is 5 um. C. Number of GFP::COSA-1 foci in meiotic nuclei at different timepoints of auxin treatment in control (blue) and pch-2 mutants (yellow). Error bars represent standard error of the mean (SEM). N values are as follows: 36 hours on auxin, control (78 nuclei) and pch-2 (83 nuclei); 24 hours on auxin, control (132 nuclei) and pch-2 (155 nuclei); 22 hours on auxin, control (152 nuclei) and pch-2 (168 nuclei); 20 hours on auxin, control (139 nuclei) and pch-2 (157 nuclei); 18 hours on auxin, control (154 nuclei) and pch-2 (154 nuclei); and 24 hours on ethanol, control (86 nuclei) and pch-2 (143 nuclei). D. Illustration of the SPO-11 depletion experiment to assay bivalents in control animals and pch-2 mutants at different timepoints of auxin treatment. Each timepoint indicates when SPO-11 is depleted in the germline with auxin induced degradation. Box indicates diakinesis, where DAPI stained bodies are analyzed. E. Oocytes from control animals and pch-2 mutants stained for DAPI (magenta). Scale bar is 4um. F. Number of DAPI stained bodies in meiotic nuclei at different timepoints of auxin treatment in control animals and pch-2 mutants. N values are as follows: 48 hours on auxin, control (47 nuclei) and pch-2 (43 nuclei); 36 hours on auxin, control (46 nuclei) and pch-2 (50 nuclei); 34 hours on auxin, control (41 nuclei) and pch-2 (41 nuclei); 32 hours on auxin, control (51 nuclei) and pch-2 (47 nuclei); 30 hours on auxin, control (46 nuclei) and pch-2 (21 nuclei); and 36 hours on ethanol, control (52 nuclei) and pch-2 (48 nuclei). Error bars represent SEM and a * indicates p-value < 0.05 and a *** indicates p-value < 0.001.

Since meiotic nuclei travel through the germline at about one cell per row per hour (Jaramillo-Lambert et al., 2007), we took advantage of this precise spatiotemporal resolution to test four different timepoints in early meiotic prophase to determine if acute depletion of SPO-11 leads to any changes in crossover formation between control animals and pch-2 mutants (Figures 3A and D). In this experiment, we assayed crossover formation with two experimental approaches. Similar to our irradiation experiment (Figures 2E and 2F), we assayed crossover formation by analyzing the number of GFP::COSA-1 foci in late pachytene 24, 22, 20 and 18 hours after acutely depleting SPO-11 (Figure 3A). In addition to this approach, we analyzed the presence of bivalents in diakinesis (Figure 3D), which indicates the successful formation of chiasmata between homologous chromosomes. Since meiotic nuclei travel from late pachytene to diakinesis in approximately 12 hours (Deshong et al., 2014), the 36, 34, 32 and 30 hour timepoints to assay bivalent formation correspond to the 24, 22, 20 and 18 hour timepoints, respectively, in which GFP::COSA-1 were analyzed. We also performed a control experiment, treating worms with ethanol for 24 and 36 hour timepoints to ensure that this tagged version of SPO-11 was fully functional for crossover formation (Figures 3C and 3F).

At the 24 hour time point (Figures 3A and 3C), when meiotic nuclei in the transition zone do not receive SPO-11-induced DSBs, we observe a range of GFP::COSA-1 foci in late pachytene in both control animals and pch-2 mutants with a majority of nuclei having 0 or 1 focus (control and pch-2 averages: 1.8 GFP::COSA-1 foci), indicating a loss of crossover formation, albeit with some heterogeneity. This heterogeneity likely reflects that it takes 1-4 hours to significantly affect DSB formation through acute degradation of SPO-11 at this stage of meiosis in C. elegans (Hicks et al., 2022). In addition, there was no statistically significant difference in the number of GFP::COSA-1 foci between control animals and pch-2 mutants. However, when we assayed bivalent formation 12 hours later (see 36 hours auxin treatment in Figures 3D and 3F), we observed the same heterogeneity and a slight, statistically significant difference in the number of DAPI stained bodies between control animals (average number of DAPI stained bodies: 10.6) and pch-2 mutants (average number of DAPI stained bodies: 9.6), suggesting that even at this very early time point, we can detect a role for PCH-2 in preventing early SPO-11-induced DSBs from becoming crossovers.

We continued to monitor GFP::COSA-1 and bivalent formation at successive timepoints. At the 22 hour and 34 hour timepoints, where SPO-11 depletion begins later in the transition zone, we saw an increase in the number of GFP::COSA-1 foci (Figure 3C) and a reduction in the number of DAPI stained bodies (Figure 3F), suggesting that more DSBs are becoming crossovers. In our analysis of GFP::COSA-1 foci at this timepoint, there appeared to a bimodal distribution: the majority of nuclei in both genotypes fell into two clear classes, those with no COSA-1 foci (24% in control animals, 26% in pch-2 mutants) and those with 6 COSA-1 foci (42% in both control animals and pch-2 mutants) (Figure 3C). The number of crossovers continued to increase at the 20 hour and 32 hour timepoints, eventually reaching wildtype numbers for GFP::COSA-1 and DAPI stained bodies by the 18 hour and 30 hour timepoints, respectively (Figures 3C and 3F).

At the 22 hour and 34 hour timepoints, we did not detect any statistically significant difference in the number of GFP::COSA-1 foci or DAPI stained bodies between control animals and pch-2 mutants (Figures 3C and 3F), even when we analyzed the number of nuclei that had greater than 6 GFP::COSA-1 foci. In contrast, while the number of GFP::COSA-1 foci was not statistically significantly different between control animals and pch-2 mutants at the 20 hour time point (Figure 3C), the number of nuclei with greater than 6 GFP::COSA-1 foci was significantly higher in pch-2 mutants (control: 3%, pch-2: 10%, p value = 0.009, Fischer’s Exact Test), similar to our analysis of GFP::COSA-1 24 hours after irradiation. Even more strikingly, the number of DAPI stained bodies was significantly lower in pch-2 animals than control animals at the corresponding 32 hour timepoint (Figure 3F), indicating that more homolog pairs had successfully formed chiasmata in pch-2 mutants. No difference in GFP::COSA-1 foci or bivalent formation was observed at the 18 hour and 30 hours timepoints, respectively (Figures 3C and 3F). These data indicate that during a narrow time frame of early meiotic prophase, likely during leptotene/zygotene given the delay in limiting DSB formation with auxin-induced degradation of SPO-11, PCH-2 prevents SPO-11-induced DSBs from becoming crossovers and chiasmata.

PCH-2 is required for timely loading and removal of MSH-5 on meiotic chromosomes through its regulation of HIM-3

After DSB formation, a subset of DSBs are licensed to be repaired through a pro-crossover pathway and eventually winnowed to a stereotypical number of crossovers (Yokoo et al., 2012). In C. elegans, these crossover-eligible intermediates can be visualized by the loading of the pro-crossover factor MSH-5, a component of the meiosis-specific MutSγcomplex that stabilizes crossover-specific DNA repair intermediates called joint molecules (Janisiw et al., 2018; Kelly et al., 2000; Snowden et al., 2004). In the spatiotemporally organized meiotic nuclei of the germline, a functional GFP tagged version of MSH-5, GFP::MSH-5, begins to form a few foci in leptotene/zygotene (the transition zone), becoming more numerous in early pachytene before decreasing in number in mid pachytene to ultimately colocalize with COSA-1 marked sites in late pachytene in a process called designation (Janisiw et al., 2018; Woglar and Villeneuve, 2018; Yokoo et al., 2012; see control in Figures 4A and 4B). The mechanism through which MSH-5 functions with the SC and other meiosis factors to ultimately form crossovers has not been established but may involve phosphorylation by cyclin-dependent kinases (Haversat et al., 2022; Zhang et al., 2021).

PCH-2 is required for timely loading and removal of MSH-5 on meiotic chromosomes through its regulation of HIM-3.

A. Representative images of nuclei in different stages of meiotic prophase in control animals and pch-2 mutants stained for DAPI (magenta) and GFP::MSH-5 (green) Scale bar is 5 um. B. Scatter plot showing average GFP::MSH-5 foci per row of germline nuclei in control animals (yellow, 163 nuclei) and pch-2 mutants (blue, 195 nuclei) from the transition zone (TZ) to late pachytene (LP), normalized to 100. The line represents a rolling average of four rows. C. Representative images of nuclei in different stages of meiotic prophase in him-3R93Y mutants (left) and pch-2;him-3R93Ydouble mutants (right), stained for DAPI (magenta) and GFP::MSH-5 (green). D. Scatter plot showing average GFP::MSH-5 foci per row in him-3R93Y(brown, 183 nuclei) and pch-2;him-3R93Y mutants (pink, 163 nuclei) from the transition zone (TZ) to late pachytene (LP), normalized to 100. The line represents a rolling average of four rows. Similar data is provided for a control germline (opaque yellow, 236 nuclei) for comparison. E. Representative images of meiotic nuclei in control animals and pch-2 mutants stained for DAPI (blue), GFP::MSH-5 (green), and OLLAS::COSA-1 (red). Yellow circles indicate GFP::MSH-5 without OLLAS::COSA-1. Scale bar is 4um. F. Scatter plot showing average GFP::MSH-5 (green) and OLLAS::COSA-1 (red) foci per row in the last five rows of the germline in control animals (36 nuclei) and pch-2 mutants (45 nuclei). The line represents a rolling average of 2 rows. G. Swarm plot showing number of GFP::MSH-5 foci in control (yellow, 14 nuclei) and pch-2 mutant (blue, 29 nuclei) nuclei with less than 6 OLLAS::COSA-1 foci (left) and 6 OLLAS::COSA-1 foci (right). Error bars represent SEM. A * indicates a p-value less than 0.05 and a ** indicates a p-value < 0.01.

Given that we have shown that PCH-2 prevents early DSBs from becoming crossovers (Figures 2 and 3), we tested whether we could cytologically detect this inhibition at the level of MSH-5 behavior in control animals and pch-2 mutants. We generated wildtype and pch-2 mutants with GFP::MSH-5 and quantified the average number of GFP::MSH-5 foci per row of nuclei from the transition zone to late pachytene (Figures 4A and 4B). We observed a statistically significant increase in the average number of GFP::MSH-5 foci per row in the transition zone in pch-2 mutants (Figures 4A and 4B, p value < 0.0001, Student’s t-test), indicating that PCH-2 typically limits MSH-5 loading at this early stage of meiotic prophase. The average number of GFP::MSH-5 foci per row were similar in number in early pachytene in both backgrounds. However, we also observed a substantial increase in the average number of GFP::MSH-5 foci per row in pch-2 mutants in the mid and late pachytene regions (Figures 4A and 4B), indicating that PCH-2 also promotes the turnover or maturation of GFP::MSH-5 foci at these later stages of meiotic prophase. Thus, PCH-2 prevents the loading of the crossover-promoting factor, MSH-5, limiting the formation of crossover-eligible intermediates during early meiotic prophase, consistent with both our genetic analysis of recombination (Figure 1) and PCH-2’s role in preventing early DSBs from becoming crossovers (Figures 2 and 3). However, PCH-2 also appears to limit the number of crossover-eligible intermediates in mid and late pachytene, an unexpected observation given that the most severe recombination defect in pch-2 mutants is the loss of crossover assurance (Deshong et al., 2014).

We previously showed that PCH-2 genetically interacts with meiotic HORMADs to control different aspects of meiosis. A mutant version of the essential meiotic HORMAD, HIM-3R93Y, binds its closure motif with reduced affinity in vitro and we have proposed that an analogous mutation in another essential meiotic HORMAD, HTP-3H96Y, behaves similarly (Russo et al., 2023). These mutations suppress different meiotic defects in pch-2 mutants in vivo, indicating that PCH-2 regulates pairing and synapsis through its regulation of HTP-3 and crossover recombination through its regulation of HIM-3 (Russo et al., 2023). To determine which meiotic HORMAD PCH-2 might be regulating to affect GFP::MSH-5’s loading and removal on meiotic chromosomes, and the potential role for these early crossover-eligible intermediates, we constructed gfp::msh-5;htp-3H96Y, gfp::msh-5;pch-2;htp-3H96Y, gfp::msh-5;him-3R93Y and gfp::msh-5;pch-2;him-3R93Y mutants and quantified GFP::MSH-5 foci throughout the germline. Both htp-3H96Y and him-3R93Ymutants showed a drastic increase in the average number of GFP::MSH-5 per row of nuclei throughout the germline (Figures 4C and 4D, Supplemental Figure 1). For example, compared to control germlines that peaked at approximately an average of 20 MSH-5 foci per row in early pachytene and then decreased in mid-pachytene, him-3R93Y germlines had closer to an average of 30 GFP::MSH-5 foci per row in early pachytene. Moreover, the average number of GFP::MSH-5 foci per row increased further in mid and late pachytene, achieving peaks closer to an average of 35 foci (Figure 4D). These very high numbers of GFP::MSH-5 were striking but do not produce an increased number of crossovers in htp-3H96Y or him-3R93Y single mutants (Russo et al., 2023), raising the possibility that not all of them reflect functional, crossover-specific intermediates, similar to what has been observed in the absence of synapsis (Woglar and Villeneuve, 2018). These data also suggest that these mutations affect the behavior of GFP::MSH-5 on meiotic chromosomes.

Quantification of GFP::MSH-5 foci in pch-2;htp-3H93Ydouble mutants was similar to htp-3H93Y single mutants, suggesting that PCH-2’s effect on the behavior of GFP::MSH-5 foci was not through its regulation of HTP-3 (Supplemental Figure 1). In stark contrast to pch-2;htp-3H93Y double mutants, the pch-2;him-3R93Y double mutant showed a dramatic decrease in the overall average of MSH-5 foci per row throughout the germline (Figures 4C and 4D). These averages fell between the averages in control and pch-2 mutant animals, particularly in the transition zone and late pachytene (Figure 4D), consistent with our previous report that this double mutant has fewer defects in crossover formation than either single mutant (Russo et al., 2023). These data indicate that PCH-2’s effect on promoting the removal of GFP::MSH-5 in both the transition zone and mid to late pachytene is through its regulation of HIM-3.

pch-2 single mutants exhibit a loss of crossover assurance, as visualized by GFP::COSA-1 foci (Figure 2B and Deshong et al., 2014). However, our quantification of GFP::MSH-5 in pch-2 single mutants in mid to late pachytene showed an increased number of crossover-eligible intermediates (Figures 4A and 4B), suggesting a complex relationship between having too many crossover-eligible intermediates and crossover assurance in C. elegans. To address this inconsistency, we generated wildtype animals and pch-2 mutants with both GFP::MSH-5 and a version of COSA-1 that has been endogenously tagged at the N-terminus with the epitope tag, OLLAS (Janisiw et al., 2018), a fusion of the E. coli OmpF protein and the mouse Langerin extracellular domain. Consistent with our and others’ previous findings, we observe a gradual reduction in the average number of GFP::MSH-5 foci per row of nuclei in control animals as these nuclei approach the end of pachytene, ultimately converging with the average number of OLLAS::COSA-1 foci per row (Figures 4E and 4F; Janisiw et al., 2018; Woglar and Villeneuve, 2018). In addition, we consistently observed co-localization of GFP::MSH-5 and OLLAS::COSA-1 foci in control animals, with a few MSH-5 foci persisting without COSA-1, consistent with crossover designation (Figures 4E and 4F).

By contrast, we detected higher average numbers of GFP::MSH-5 foci per row across all late pachytene nuclei in pch-2 mutants and this average never converged upon the average number of OLLAS::COSA-1 foci per row (Figures 4E and 4F). To test whether there was a correlation between the number of GFP::MSH-5 foci and the loss of crossover assurance in pch-2 mutants, we determined the average number of GFP::MSH-5 foci in meiotic nuclei with 6 OLLAS::COSA-1 foci and those with less than 6 OLLAS::COSA-1 foci in both control and pch-2 mutant animals (Figures 4E and 4G, Supplemental Figure 2). Indeed, we observed that pch-2 mutant nuclei with less than 6 OLLAS::COSA-1 foci had significantly higher numbers of GFP::MSH-5 foci, compared to both control and pch-2 mutant nuclei with 6 OLLAS::COSA-1 foci (Figures 4E and 4G), suggesting that the inability to reduce the number of crossover-eligible intermediates in pch-2 mutants, counterintuitively, prevents designation on some chromosomes and contributes to the observed loss of crossover assurance.

PCH-2 is removed from the synaptonemal complex when crossovers are designated

We have previously shown that PCH-2 localization to the SC is extended when there are partial defects in synapsis or changes in karyotype, producing an increase in crossovers and a loss of crossover interference (Deshong et al., 2014; Patel et al., 2023). These data led us to hypothesize that PCH-2’s presence on chromosomes promotes crossover formation. However, our quantification of GFP::MSH-5 indicates that PCH-2 is required to limit the number of crossover-eligible intermediates, in direct contrast to our proposed hypothesis (Figure 4). Therefore, we decided to revisit what role PCH-2 localization to the SC might play in regulating crossover formation. To this end, we localized PCH-2 in dsb-2 mutants, in which DSB formation is substantially reduced and fewer crossovers are formed in C. elegans (Rosu et al., 2013). We performed this experiment because, in budding yeast, similar mutants that reduce DSB formation rely on Pch2 to successfully complete meiosis (Joshi et al., 2009; Zanders and Alani, 2009).

Because the recombination defect in dsb-2 mutants worsens with age (Rosu et al., 2013), we looked at PCH-2 localization in both young (24 hours post L4 larval stage) and older animals (48 hours post L4). We observed a unique and striking localization pattern in dsb-2 mutants that we had not observed before. When stained for PCH-2 and GFP::COSA-1, nuclei retain PCH-2 onto chromosomes far into late pachytene, past the normal region when PCH-2 typically is removed from chromosomes (Figures 5A and 5B). However, unlike what we have observed in other mutants that have defects in synapsis or changes in karyotype (Deshong et al., 2014; Patel et al., 2023), this retention of PCH-2 is not uniform among all late pachytene nuclei in dsb-2 mutants. Instead, most nuclei lose PCH-2 localization in late pachytene while some retain it. We tested whether there was a relationship between the retention of PCH-2 and the number of GFP::COSA-1 foci and found that 96% of nuclei that lose PCH-2 have at least 1 GFP::COSA-1 focus (Figures 5A and 5B). In older animals (see 48 hours post-L4), this pattern was even more clear: 77% of nuclei without PCH-2 had 1 or more GFP::COSA-1 focus and 93% of meiotic nuclei without a GFP::COSA-1 focus retained PCH-2.

PCH-2 is removed when crossovers are designated.

A. Representative images of meiotic nuclei in dsb-2 animals 24 hours post L4 and 48 hours post L4 stained for DAPI (magenta), PCH-2 (red), and GFP::COSA-1 (green). Scale bar is 4 um. B. Stacked histograms showing percentage of PCH-2 positive (n = 43 at 24 hours, n = 42 at 48 hours) and negative (n = 194 at 24 hours, n = 130 at 48 hours) nuclei with (lime) and without (dark green) GFP::COSA-1 foci in dsb-2 mutants at 24 hours post L4 and 48 hours post L4. C. Representative images of meiotic nuclei in dsb-2::AID and dsb-2::AID;pch-2 mutants treated with auxin and stained for DAPI (magenta) and GFP::COSA-1 (green). Scale bar is 5 um. D. Swarm plot showing the number of GFP::COSA-1 foci in dsb-2::AID (gray) and dsb-2::AID;pch-2 (lemon) mutants when treated with ethanol or auxin. N values are as follows: dsb-2::AID (101 nuclei) and dsb-2::AID;pch-2 (154 nuclei) on ethanol, dsb-2::AID (136 nuclei) and dsb-2::AID;pch-2 (131 nuclei) on auxin. Error bars represent the SEM. A ** indicates a p-value less than 0.01 and a **** indicates p-value < 0.0001.

These data indicate that PCH-2’s removal from the SC is in response to, coincident with or facilitates crossover designation. To distinguish between these possibilities, we analyzed GFP::COSA-1 foci in dsb-2::AID (Zhang et al., 2018) and dsb-2::AID;pch-2 worms, reasoning that if PCH-2’s removal is in response to or coincident with designation, we should not detect any differences in the number of GFP::COSA-1 foci in dsb-2::AID and dsb-2::AID;pch-2 worms. We performed these experiments with the auxin-inducible degradation system because of variability in the dsb-2 mutant background that affected reproducibility of our experiments.

Upon auxin treatment of dsb-2::AID worms, we observed a statistically significant (p value < 0.0001, Mann-Whitney U test) decrease in the number of GFP::COSA-1 foci, in comparison with ethanol-treated worms, verifying that we can reliably knock down DSB-2 with the AID system (Figure 5D). When we performed the same experiment in dsb-2::AID;pch-2 worms, the average number of GFP::COSA-1 foci further decreased, indicating that fewer crossovers are designated when DSBs are reduced and PCH-2 is absent (Figures 5C and 5D). These data argue against the possibility that PCH-2’s removal from the SC is simply in response to or coincident with crossover designation and instead, suggest that PCH-2’s removal from the SC somehow facilitates crossover designation and assurance. Given the correlation between elevated GFP::MSH-5 foci and the loss of crossover assurance we observe in pch-2 mutants, we propose that PCH-2 is retained on meiotic chromosomes to ensure that extra crossover-eligible intermediates are removed and crossover designation is delayed until crossover assurance can be guaranteed in C. elegans.

PCH-2 and high CHK-2 activity control the fate of early double strand breaks

In C. elegans, meiotic cell cycle entry and progression depends on the activity of CHK-2, the meiosis-specific ortholog of the DNA-damage kinase Chk2/CHEK2 (Baudrimont et al., 2022; Castellano-Pozo et al., 2020; Kim et al., 2015; MacQueen and Villeneuve, 2001; Zhang et al., 2023). Meiotic nuclei in leptotene/zygotene are characterized by high CHK-2 activity, which drops to intermediate activity in mid-pachytene (Kim et al., 2015; Zhang et al., 2023). In late pachytene, CHK-2 activity is inactivated by the recruitment of polo-like kinase, PLK-2, to the SC, which enables crossover designation (Zhang et al., 2023). Thus, CHK-2 activity also has consequences for the progression of meiotic recombination and DNA repair.

Given that early DSBs are prevented from becoming crossovers by PCH-2 in early meiotic prophase, we wanted to test if high CHK-2 activity in leptotene/zygotene also contributed to the fate of these early DSBs. To evaluate this possibility, we used syp-1T452A mutants (Sato-Carlton et al., 2018). The SC component, SYP-1, is phosphorylated by cell cycle kinase, CDK-1, on T452, producing a Polo box binding motif that recruits the polo-like kinase, PLK-2, to the SC, contributing to the inactivation of CHK-2 when chromosomes are synapsed (Brandt et al., 2020; Zhang et al., 2023). Therefore, in syp-1T452A mutants, CHK-2 activity remains high throughout most of meiotic prophase in the C. elegans germline, delaying meiotic progression as visualized by the extension of the transition zone (Figure 6A, Supplemental Figure 3).

PCH-2 and high CHK-2 activity control the fate of early double strand breaks.

A. Illustration of CHK-2 activity in wildtype and syp-1T452A germlines. B. Representative images of meiotic nuclei late pachytene in syp-1T452A and pch-2;syp-1T452Amutants stained for DAPI (magenta) and GFP::COSA-1 (green). Scale bar is 5 um. C. Swarm plot showing number of GFP::COSA-1 foci in control animals (blue), pch-2 (yellow), syp-1T452A (maroon), and pch-2;syp-1T452A(light blue) mutants. Error bars represent SEM. D. Oocytes from syp-1T452Aand pch-2;syp-1T452A mutant worms stained for DAPI (magenta). Scale bar is 4um. D. Swarm plot showing number of DAPI stained bodies in control animals (blue, n = 154), pch-2 (yellow, n = 89), syp-1T452A (maroon, n = 247)), and pch-2;syp-1T452A(light blue, n = 242) mutants. Error bars represent SEM. A **** indicates p-value < 0.0001.

We quantified the total number of GFP::COSA-1 foci in late pachytene nuclei in syp-1T452A mutants and observed a significant decrease in the average number of GFP::COSA-1 foci when compared to control animals (Figures 6B and 6C), consistent with previous findings that syp-1T452A mutants delay designation and have fewer crossovers (Zhang et al., 2023). These data are also consistent with the possibility that some DSBs fail to become crossovers when CHK-2 activity remains high. If this hypothesis is correct, we predict that the combination of high CHK-2 activity and loss of pch-2 should produce more crossovers. We generated pch-2;syp-1T452A double mutants to test this hypothesis. These double mutants exhibited the same delay in meiotic progression as syp-1T452A single mutants (Supplemental Figure 3). When we quantified GFP::COSA-1 in pch-2;syp-1T452A double mutants, we detected a significant increase in the average number of GFP::COSA-1 foci compared to syp-1T452Asingle mutants (Figure 6C), in strong support of our hypothesis.

To further verify that the pch-2 mutation suppresses the crossover defect in syp-1T452A mutants, we also monitored bivalent formation in diakinesis nuclei. syp-1T452A single mutants exhibit an average of 7.12 DAPI staining bodies in diakinesis, higher than both control animals and pch-2 single mutants (Figure 6D). By contrast, pch-2;syp-1T452A mutants had an average of 6.05 DAPI staining bodies (Figure 6D and 6E), directly supporting our COSA-1 analysis and indicating that both PCH-2 function and high CHK-2 activity collaborate to control the fate of DSBs and prevent some of them from becoming crossovers in early meiotic prophase.

Discussion

We have shown that PCH-2 antagonizes crossover formation throughout meiotic prophase (Figures 2, 3, and 4) and that this regulation occurs through one of the three essential meiotic HORMADs in C. elegans, HIM-3 (Figure 4). We propose that PCH-2 remodels HIM-3 on meiotic chromosomes to destabilize crossover-eligible intermediates, visualized in our experiments as GFP-MSH-5 foci, thus limiting which DSBs will become crossovers (Figure 7). However, this antagonism has different consequences depending on when during meiotic prophase it occurs, underscoring the importance of temporal regulation of these events. During leptotene/zygotene, when CHK-2 activity is high (Kim et al., 2015; Zhang et al., 2023) and PCH-2 is present as foci on chromosomes (Deshong et al., 2014), PCH-2 prevents crossover formation at some sites of initial DSB formation and early homolog interactions (Figure 7B). In this way, PCH-2 promotes a wider distribution of crossovers across the genome (Figure 1). In pachytene, when CHK-2 activity has decreased (Kim et al., 2015; Zhang et al., 2023) and PCH-2 is localized to the SC (Deshong et al., 2014), PCH-2 winnows crossover-eligible intermediates marked by GFP-MSH-5, ensuring their designation, colocalization with COSA-1 and crossover assurance (Figure 7B). When there are defects in recombination, such as partial synapsis (Deshong et al., 2014), changes in karyotype (Patel et al., 2023) or too few DSBs (Figure 5), PCH-2 persists on the SC to prevent designation, guarantee crossover assurance and some degree of homeostasis, independent of an additional feedback mechanism that increases DSB formation (Patel et al., 2023). That persistence of PCH-2 also disrupts crossover interference in two of these situations (Deshong et al., 2014; Patel et al., 2023) strongly suggests that crossover interference is mechanistically linked to assurance and homeostasis. For example, PCH-2’s persistence on the synaptonemal complex may maintain an extended period of competency for interhomolog repair, as has been observed in mutants defective in crossover recombination (Rosu et al., 2011).

PCH-2 remodels HIM-3 to disassemble crossover-eligible intermediates, controlling crossover distribution and number.

A. Model for how pch-2 and him-3R93Ymutations genetically interact to affect the progression of meiotic recombination. HIM-3 adopts the closed conformation upon binding an interacting protein with a closure motif and its conversion to the extended conformation is facilitated by PCH-2’s remodeling of its HORMA domain. B. Model for how PCH-2 and HIM-3 progressively implement meiotic recombination during different stages of meiotic prophase.

him-3R93Y;pch-2 double mutants have stronger crossover assurance than either single mutant but less than wildtype animals (Russo et al., 2023), a phenotype which can now be explained by the behavior of GFP::MSH-5 foci in these double mutants (Figures 4C and 4D). We have shown that HIM-3R93Y mutant protein can adopt the closed conformation and loads on meiotic chromosomes similar to wildtype HIM-3 (Russo et al., 2023). In vitro analysis shows that HIM-3R93Y binds its closure motif with reduced affinity, likely affecting its ability to adopt the closed conformation in vivo (Figure 7A and Russo et al., 2023). The proposed role of PCH-2 in meiosis is to remodel meiotic HORMADs from the closed conformation to the extended one (Figure 7A), regulating their association with chromosomes (Bhalla, 2023). However, since meiotic HORMADs are not visibly depleted from chromosomes during meiotic progression in C. elegans (Couteau et al., 2004; Couteau and Zetka, 2005; Goodyer et al., 2008; Martinez-Perez and Villeneuve, 2005), this genetic interaction supports a role for PCH-2 in temporarily reducing the occupancy of meiotic HORMADs on meiotic chromosomes, destabilizing interactions with partner proteins that modulate the progression and fidelity of meiotic recombination (Russo et al., 2023). Thus, PCH-2’s remodeling of HIM-3 would disrupt protein-protein interactions that underlie crossover-eligible intermediates, destabilizing them and reducing their number on chromosomes, contributing to crossover control (Figure 7B). Since meiotic HORMADs are essential for meiotic chromosome axis structure and function, our data therefore support a role for the meiotic axis in crossover control, as suggested by previous reports (Chu et al., 2024; Girard et al., 2023; Lambing et al., 2020; Nabeshima et al., 2004). However, the disassembly of crossover-eligible intermediates would also release pro-crossover factors present at these sites, such as MSH-5, to concentrate at other, more stable sites on the SC, facilitating designation during synapsis (Girard et al., 2023; Yokoo et al., 2012). In this way, PCH-2’s mechanism of action may reconcile current competing models of crossover control (Girard et al., 2023). Moreover, these results raise the possibility that this behavior of PCH-2 and/or meiotic HORMADs might be regulated by post-translational modifications associated with crossover control, such as ubiquitination, SUMOylation and phosphorylation (Gray and Cohen, 2016).

Given that mutation of PCH-2 produces changes in the number and distribution of crossovers across multiple model systems, we argue that controlling crossover distribution and number is the conserved role of PCH-2 in meiosis (Bhalla, 2023). We have previously proposed that the conserved role of PCH-2 is to coordinate meiotic recombination with synapsis (Bhalla, 2023). With these results, we further refine this model. Early in meiotic prophase, PCH-2 remodels meiotic HORMADs to prevent some DSBs from becoming crossover-eligible intermediates, widening the recombination landscape beyond early homolog interactions and/or sites that tend to be more favorable for DSB formation, also known as “hot spots.” For example, this may explain the localization of TRIP13, the PCH-2 ortholog in mice, to telomeres (Chotiner et al., 2024), sites that experience early homolog interactions due to the organization of meiotic chromosomes in the bouquet formation (Scherthan et al., 1996). This antagonism may also expand the regions of the genome that initiate synapsis in organisms that use DSBs to accomplish this event. This possibility is supported by the observation that loss of TRIP13 in mammals produce meiotic chromosomes that exhibit partial asynapsis (Roig et al., 2010), particularly near regions that may act as barriers to SC polymerization (Brown et al., 2005; Roig et al., 2010). Limiting which DSBs becomes crossover-eligible intermediates in early meiotic prophase also ensures that meiotic recombination overlaps with synapsis, either completely, as in C. elegans (Yokoo et al., 2012) or partially, as in budding yeast, plants and mice (Capilla-Perez et al., 2021; Cole et al., 2012; Joshi et al., 2015; Morgan et al., 2021). Once synapsis is complete, PCH-2 continues to remodel meiotic HORMADs on chromosomes to control the gradual implementation of crossover number and distribution, reinforcing the important role that synapsis plays in mediating crossover control (Durand et al., 2022; Libuda et al., 2013).

Unexpectedly, the inability to reduce the number of crossover-eligible intermediates in pch-2 mutants, as visualized by GFP::MSH-5 foci, does not produce extra crossovers but a loss of crossover-assurance in C. elegans (Figure 4), a somewhat counterintuitive result. One interpretation of these data is that crossover-eligible intermediates may be more numerous but absent from some chromosomes in pch-2 mutants, explaining the loss of crossover assurance. Since the absence of crossover intermediates in C. elegans is accompanied by premature desynapsis of individual chromosomes (Machovina et al., 2016; Pattabiraman et al., 2017) and chromosomes in pch-2 mutants delay desynapsis (Deshong et al., 2014), we do not favor this interpretation. Instead, we propose that having too many crossover-eligible intermediates can be as deleterious to crossover assurance as having too few (Figure 7B). This possibility is further supported by the loss of crossover assurance we detect in irradiated wildtype worms, which is exacerbated in pch-2 mutants (Figure 2).

This phenomenon, where crossover-eligible intermediates need to be winnowed to some threshold number to ensure crossover assurance, may explain the loss of crossover assurance also observed in Trip13 deficient mice (Roig et al., 2010) and on small chromosomes in budding yeast (Chakraborty et al., 2017). Alternatively, the counterintuitive relationship between the number of crossover-eligible precursors and crossover assurance in pch-2 mutants we observe might reflect an additional layer of regulation during crossover formation specific to C. elegans. Since C. elegans chromosomes are holocentric, crossovers play an additional role organizing chromosomes for the ordered release of sister chromatid cohesion during meiosis I (Martinez-Perez et al., 2008; Nabeshima et al., 2005) and extra crossovers can be deleterious to accurate chromosome segregation (Hollis et al., 2020). By contrast, in Arabidopsis, a system that appears to be able to tolerate an extraordinarily high number of crossovers with little to no effect on chromosome segregation (Durand et al., 2022), PCH2’s inability to localize to the SC produces an increase in crossover formation (Figure 7), as visualized by both MLH1 foci and the formation of chiasmata (Yang et al., 2022). Once again, an overarching theme that becomes apparent in our model is that PCH-2 may play a common role in different systems, with dramatic variations in phenotypic consequences given species-specific requirements and constraints.

We were not surprised to see high numbers of double crossovers on almost every chromosome in our genetic analysis of recombination in wildtype worms, given our previous analysis (Deshong et al., 2014). However, we were surprised to see that the majority of them were found near Pairing Centers and sites of synapsis initiation, suggesting a relationship between early homolog interactions and the formation of double crossovers. When we revisited our previous data, we observed similar patterns. In addition, we do not detect these double crossovers cytologically in C. elegans, even when using the OLLAS::COSA-1 reporter, which has been reported to identify double crossovers in spermatogenesis not visualized by GFP::COSA-1 (Cahoon et al., 2023). Crossovers that are cytologically marked by COSA-1 are known as Class I crossovers, which depend on pro-crossover factors such as MSH-5 and ZHP-3, rely on synapsis and exhibit crossover control (Gray and Cohen, 2016). These data raise the intriguing possibility that these double crossovers are the product of the alternate, Class II, pathway of crossover formation, which relies on a different suite of proteins, does not respond to crossover control, do not depend on synapsis and contributes to varying degrees in different model systems (Gray and Cohen, 2016; Youds et al., 2010). Thus, based on the close, functional relationship that exists between Class I crossovers and synapsis and the apparent antagonistic relationship that exists between Class II crossovers and synapsis, important corollaries of our model may be that PCH-2 specifically coordinates recombination with synapsis to promote Class I crossovers, limit Class II crossovers and that Class II crossovers are more likely to form early in meiosis, prior to synapsis. Therefore, variations in the contribution of the Class II crossover pathways to crossover recombination and the degree of cross-talk between Class I and Class II pathways among model systems might reflect the degree to which crossover formation overlaps with synapsis (Bhalla, 2023; Gray and Cohen, 2016; Yokoo et al., 2012). Furthermore, these corollaries are entirely consistent with PCH-2’s absence from the genome of fission yeast and Tetrahymena (Kops et al., 2020; Wu and Burgess, 2006), both systems in which chromosomes do not undergo meiotic synapsis, crossovers do not exhibit interference and all crossovers are dependent on the Class II pathway (Hollingsworth and Brill, 2004; Lukaszewicz et al., 2013; Wolfe et al., 1976). However, this aspect of the model needs to be formally tested.

Our results also have some additional important implications about the regulation of DSB formation across the genome in C. elegans. It is formally possible that PCH-2 controls DSB distribution. However, the complexity of recombination defects in pch-2 mutants argues against this possibility. Instead, we explain the shift in the recombination landscape away from the central regions of chromosomes and toward PC ends in pch-2 mutants (Figure 1) as a result of early DSBs becoming crossovers at the expense of later DSBs. This explanation is better supported by the panel of defects observed in pch-2 mutants: DSBs introduced early in meiosis become crossover eligible intermediates and crossovers (Figures 2, 3 and 4) and while the number of DSBs are constant, DNA repair is accelerated (Deshong et al., 2014). Moreover, this explanation suggests that when DSBs happen in meiotic prophase affects where they happen in the genome. Specifically, we propose that chromosome arms, which are gene poor, receive DSBs early during (or even throughout) DSB formation and the center of chromosomes, which are gene rich, receive DSBs later. The shift in recombination to the center of chromosomes when defects in meiosis prolong DSB formation provide further support to this possibility (Carlton et al., 2006; Deshong et al., 2014). A similar regulation of the timing of DSB formation has been demonstrated in budding yeast, where the DSB landscape across the whole genome expands when time in prophase is increased (Lopez Ruiz et al., 2024) and small, highly recombinogenic, chromosomes, get more DSBs later in meiotic prophase (Murakami et al., 2020; Subramanian et al., 2019). However, this temporal regulation has not been reported previously in C. elegans and suggests that this phenomenon is more widely conserved. This expansion of the DSB landscape in C. elegans to include the center regions of chromosomes later in meiosis may be a deliberate attempt for recombination to create new haplotypes for evolution to act on, despite the relative paucity of DSBs and the observation that they can result in chromosome missegregation (Altendorfer et al., 2020).

Finally, our work raises some important questions about the functional role(s) of DSBs in meiosis, aside from their contributions to crossover formation. Hicks and colleagues were the first to report that early DSBs do not contribute to crossover formation in C. elegans (Hicks et al., 2022). Here we show that these early DSBs are prevented from becoming crossovers by both PCH-2 activity and cell cycle stage, specifically in leptotene/zygotene, when homologs are initiating pairing and synapsis. In budding yeast, similar, early-occurring DSBs have been characterized as “scout DSBs” because of their preference for repair from sister chromatids, versus homologous chromosomes, and their proposed role in contributing to homolog pairing (Borde and de Massy, 2015; Joshi et al., 2015). The homolog bias that these “scout DSBs” do display seems dependent on budding yeast PCH2 (Joshi et al., 2015) but interpreting this experiment is complicated by the fact that Pch2 is also required to make the budding yeast meiotic HORMAD, Hop1, available for its loading onto meiotic chromosomes (Herruzo et al., 2021).

In contrast to budding yeast, C. elegans does not rely on DSBs to promote homolog pairing and initiate synapsis (Dernburg et al., 1998); in worms, cis-acting sites called Pairing Centers are essential for homolog pairing and synapsis (MacQueen et al., 2005). There is certainly some support in the literature for DSBs playing a role in supporting pairing and synapsis in C. elegans (Guo et al., 2022; Mlynarczyk-Evans et al., 2013; Roelens et al., 2015). However, if these early DSBs were contributing to pairing and synapsis, we would expect to see a genetic interaction between htp-3H96Y and pch-2 mutations in the installation of GFP::MSH-5 in the transition zone; we previously reported that htp-3H96Y suppresses the acceleration of pairing and synapsis of pch-2 mutants, particularly when Pairing Center function is compromised (Russo et al., 2023). Instead, we favor the possibility that these early DSBs are generated to amplify the signaling of DNA damage kinases to prime them for their role in recombination, a role that has also been proposed for “scout” DSBs (Joshi et al., 2015). That ATM-1, a conserved DNA damage kinase that is downstream of CHK-2 in C. elegans, relies on DSBs for full activity, supports this proposed, conserved role (Yu et al., 2023).

Our work provides an important framework to finally understand the role of PCH-2 in controlling the number and distribution of crossovers, a role that we argue is its conserved role. While specific details may vary across systems, we propose that PCH-2 remodels meiotic HORMADs throughout meiotic prophase to destabilize crossover-eligible precursors, coordinating meiotic recombination with synapsis, contributing to the progressive implementation of meiotic recombination and guaranteeing crossover assurance, interference and homeostasis.

Methods and materials

C.elegans Genetics and Genome Engineering

The C. elegans Bristol N2 was used as the wild-type strain. All strains were maintained at 20°C under standard conditions unless stated. Mutant combinations were generated by crossing. The following mutants and rearrangements were used:

LGII: pch-2(tm1458), meIs8 ([pie-1p::GFP::cosa-1 + unc-119(+)], dsb-2(me96), dsb-2(ie58[dsb-2::AID::3xFLAG])

LGIII: cosa-1(ddr12[OLLAS::cosa-1]); htp-3(vc75)

LGIV: him-3(blt9), spo-11(ie59[spo-11::AID::3xFLAG]), msh-5[ddr22(GFP::msh-5)], ieSi38 [sun-1p::TIR1::mRuby::sun-1 3’UTR + Cbr-unc-119(+)]

LGV: syp-1(icm85[T452A]), nT1[qIs51], bcIs39 (Plin-15::ced-1::GFP)

Genetic analysis of Recombination

The wildtype Hawaiian CB4856 strain (HI) and the Bristol N2 strain were used to assay recombination between single nucleotide polymorphisms (SNPs) on Chromosomes I, III, IV and X (Bazan and Hillers, 2011; Wicks et al., 2001). The SNPs, primers, enzymes used for restriction digests and expected fragment sizes are included in Supplemental Table 1. To measure wild-type recombination, N2 males containing bcIs39 were crossed to Hawaiian CB4856 worms. Cross-progeny hermaphrodites were identified by the presence of bcIs39 and contained one N2 and one CB4856 chromosome. These were assayed for recombination by crossing with CB4856 males containing myo-2::mCherry. Cross-progeny hermaphrodites from the resulting cross were isolated as L4s, and then cultured individually in 96-well plates in liquid S-media complete supplemented with HB101. Four days after initial culturing, starved populations were lysed and used for PCR and restriction digest to detect N2 and CB4856 SNP alleles.

For recombination in pch-2 mutants, strains homozygous for the CB4856 background of the relevant SNPs were created by backcrossing pch-2 mutants to worms of the CB4856 background at least eight times and verifying the presence of Hawaiian SNPs on all chromosomes tested in the recombination assay. These Hawaiianized pch-2 mutants were then mated with pch-2; bcIs39. Subsequent steps were performed as in the wild-type worms.

Immunostaining

DAPI staining and immunostaining was performed as in (Russo et al., 2023), 20 to 24 hours post L4 unless otherwise noted. For analyzing bivalents, the same protocol was implemented with the exception that hermaphrodites were dissected and DAPI stained 48 hours post late L4 stage, unless otherwise noted.

The following primary antibodies were used at the indicated dilutions: alpaca anti-GFP Booster (ChromoTek, gb2AF488) was used at 1:1000; rat anti-OLLAS (Invitrogen, PIMA516125) was used at 1:1000; and rabbit anti PCH-2 (Deshong et al., 2014) was used at 1:500. The following secondary antibodies were used at the indicated dilutions: anti-rabbit Cy3 (Jackson Labs) was used at 1:500 and anti-rat Cy5 (Jackson Labs) was used at 1:500

Irradiation Experiments

Control and pch-2 mutant L4’s were aged 12-14 hours before being exposed to 1,000 rad (10 Gy) of X-ray radiation using a Precision MultiRad 160 X-irradiator (Precision X-Ray Inc.). Germlines were then fixed and stained, 8 hours and 24 hours post irradiation.

Auxin-induced Degradation Experiments

Auxin treatment was performed by transferring young adult worms (aged 12-14 hours post-L4) to bacteria-seeded plates containing auxin or 99% ethanol at specific time points, except for experiments in which L4s were transferred directly to bacteria-seeded plates containing auxin or ethanol (the 36 hour and 48 hour time points in Figure 3C and E and the experiments in Figures 5C and D). The natural auxin indole3-acetic acid (IAA) was purchased from Alfa Aesar (#A10556). A 400 mM stock solution in ethanol was prepared and was stored at 4°C for up to one month. Auxin was diluted to 100mM, and 100ul was spread onto NGM plates. Plates were allowed to dry before seeding with fresh OP50 culture. Plates were left at 20°C for 2-3 days in the dark to allow for bacterial lawn growth.

Imaging and Quantification

All images were acquired using a DeltaVision Personal DV system (Applied Precision) equipped with a 100X N.A. 1.40 oil-immersion objective (Olympus), resulting in an effective XY pixel spacing of 0.064 or 0.040 µm. Three-dimensional image stacks were collected at 0.2-µm Z spacing and processed by constrained, iterative deconvolution. Image scaling, analysis and maximum-intensity projections were performed using functions in the softWoRx software package.

For analysis of GFP::MSH-5 foci and meiotic progression, sum projections were generated using ImageJ for each image of the germline. ImageJ plugins Cell Counter, ROI Manager, and Find Maxima were used to identify and quantify GFP::MSH-5 foci by row from transition zone to the end of pachytene. The threshold value was set depending on background conditions to ensure minimal signals were identified. Foci were only quantified if they co-localized with DAPI staining. For all genotypes, three germlines per genotype were analyzed and representative germlines are shown.

Graphing and Statistical Analysis

Data was analyzed using Python 3.8 and Prism for statistical significance. All datasets were tested for normality using the Shapiro-Wilk test. For Figures 1, 2 and 5B, Fisher’s exact test was used to determine significance. For Figures 3, 4, 5D and 6, Mann-Whitney U test was used to determine significance.